---
title: 1. Building larger Rcpp programs
weight: 1
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>


### Spreading `Rcpp` code over multiple `C++` files

So far in these notes, we have put all our `Rcpp` code in a single `.cpp` file or we have written the `Rcpp` code directly in `R`. An example of the latter approach is:
```{r, eval = FALSE}
library(Rcpp)
sourceCpp(code = '
#include <Rcpp.h>
using namespace Rcpp;

double mySquare(double x){
  return x * x;
}

// [[Rcpp::export(name = "squareVect")]]
NumericVector squareVect_I(NumericVector v) {
  
  NumericVector out( v.length() );
  for(int ii = 0; ii < v.length(); ii++){
    out[ii] = mySquare(v[ii]);
  }
  return out;
  
}')
```
where the `squareVect_I` `C++` function simply applies the `mySquare` function to each element of the vector `v`. When building larger programs, it is often convenient to spread the code over several source files, hence here we explain how to do it using header file.

Suppose that we have a source code file, `squareVect.cpp`, containing the function that we want to export to `R`:
```{r engine = 'Rcpp', eval = FALSE}
#include <Rcpp.h>
#include "utilities.h"

using namespace Rcpp;

// [[Rcpp::export(name = "squareVect")]]
NumericVector squareVect_I(NumericVector v) {
  
  NumericVector out( v.length() );
  for(int ii = 0; ii < v.length(); ii++){
    out[ii] = mySquare(v[ii]);
  }
  return out;
  
}
```
Notice that here we are including the `utilities.h` header file, which contains:
```{r engine = 'Rcpp', eval = FALSE}
#ifndef __UTILITIES__
#define __UTILITIES__

double mySquare(double x);

#endif // __UTILITIES__
```
Here `#ifndef`, `#define` and `#endif` are directives that will be used by the preprocessor to do some code editing before compilation. In particular, we define `__UTILITIES__` and then we add the `#ifndef __UTILITIES__` directive to avoid including the same header file twice by mistake. The line `double mySquare(double x);` is a declaration of the `mySquare` function, simply stating its name, and the type of its inputs and output. The definition of the function is contained in `utilities.cpp`: 
```{r engine = 'Rcpp', eval = FALSE}
#include "utilities.h"

double mySquare(double x){
  return x * x;
}
```
Now we compile the code using `sourceCpp`:
```{r}
library(Rcpp)
sourceCpp(file = "squareVect.cpp")

squareVect(1:5) 
```
which seems to work fine.

### Using `Rcpp` extensions with `sourceCpp`

We have seen in a previous chapter that `sourceCpp` and `cppFunction` take care of the compiler flags and options needed to compile and link `Rcpp` programs, while such flags need to be handled manually when using `R CMD SHLIB`. For example, if we compile the following `Rcpp` function:
```{r, output.lines = 1:30}
sourceCpp(code = '
 #include <Rcpp.h>
 using namespace Rcpp;

 // [[Rcpp::export(name = "exampleCpp")]]
 NumericVector exampleCpp_I() {
   NumericVector out(10);
   return out;
}')
```
then `R CMD SHLIB` will be use the option `-I"some_directory/Rcpp/include"` when compiling, which allows the compiler to find `Rcpp.h`. Similarly, when we use `RcppArmadillo` in our `Rcpp` code, as in this example:
```{r}
sourceCpp(code = '
 // [[Rcpp::depends(RcppArmadillo)]]
 #include <RcppArmadillo.h>
 using namespace Rcpp;

 // [[Rcpp::export(name = "MMv_arma")]]
 arma::vec Mv_I(arma::mat& A, arma::vec& y) {
   return A * y;
}')
```
we include the `// [[Rcpp::depends(RcppArmadillo)]]` attribute, to make sure that the compiler options are set correctly to work with `Armadillo`. In particular, in this case `R CMD SHLIB` will need (among others) the option `-I"some_directory/RcppArmadillo/include"` to find `RcppArmadillo.h` as well as `-llapack` and `-lblas` to include the `LAPACK` and `BLAS` linear algebra libraries, which are used by `Armadillo`. You can see this by setting `verbose = TRUE` when calling `sourceCpp`.

The point is that the `Rcpp::depends` mechanism gives you access to lots of extensions that can be used within your `Rcpp` code, without having to worry about the compilation process. In particular, in addition to `RcppArmadillo`, you can use:

   - `RcppEigen` which provides access to the `Eigen` linear algebra library;
   - `BH` which provides access to the `Boost` `C++` library;
   - `RcppGLS` for integration with `GNU Scientific Library` (or `GSL`) library 
      of numerical routines;
   - `RcppMLPACK` for integration with the `MLPACK` `C++` machine learning library;
   - `RcppParallel` for parallel computation in `Rcpp`;
   - and several other packages...
   
For example, the following code use the RNG provided by the `Boost` `C++` library:
```{r}
library(Rcpp)
sourceCpp(code = '
 // [[Rcpp::depends(BH)]]
 #include <Rcpp.h>
 #include <random>
 #include <boost/random/normal_distribution.hpp>
 using namespace Rcpp;

 // [[Rcpp::export(name = "rnormBoost")]]
 NumericVector rnorm_I(int n, int seed) {
 
  // Set up Marsenned Twister RNG from the C++ Standard Library
  // and initialize is at a given seed
  std::mt19937 eng;
  eng.seed(seed);
  
  // Create a standard normal distribution using Boost
  boost::normal_distribution<> normal(0.0, 1.0);
  
  NumericVector out(n);
  
  for(int ii = 0; ii < n; ii++){
   // Pass the RNG to the normal distribution to sample
   out[ii] = normal(eng); 
  }
  
  return out;

}')

rnormBoost(5, 3)
rnormBoost(6, 3)
rnormBoost(6, 4)
```

The packages listed above are some of the packages on which your `Rcpp` code can depend via the `Rcpp::depends` attribute. Understanding exactly how the latter works is beyond the scope of this course, but it is useful to be aware of the fact that `sourceCpp` will look for the information on the package(s) mentioned in `Rcpp::depends`, which is needed to set up the compilation enviroment. For example, it will look for the path to the header file to be included, which can be found by doing:
```{r}
system.file("include", package = "RcppArmadillo")
```
Further information can retrieved via the `getPlugin` function, for example:
```{r}
get("inlineCxxPlugin", asNamespace("RcppArmadillo"))()$env
```
gives the additional `RcppArmadillo` environment variables required to successfully compile code depending on `Rcpp`. Again, we don't need to know exactly how this works, but it is good to be aware of the fact that `sourceCpp` is doing some work to set up the compilation environment. Other `Rcpp` functions, such as `cppFunction`, work similarly.

In this section we have seen how to you can spread your `Rcpp` code over multiple files to build larger applications. However, a better approach to manage and distribute your `Rcpp` code is building an `R` package. The next section will explain how to do it.




