---
title: 3. Using Rcpp in an R package (2)
weight: 3
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

```{r, echo = FALSE, message = FALSE}
library(Rcpp)
system("rm -r mypack")
system("rm -r secondPack")
Rcpp.package.skeleton("mypack")
```

### Using `C++` code from other packages within your package

In the previous section we explained how to build a basic `R` package containing `Rcpp`-based `C++` code. Here we explain how you can use `C++` code from other packages (e.g., `RcppArmadillo`) in your package.

Consider the following function, which calculates the dot-product between two vectors using `RcppArmadillo`:
```{r, comment=NA}
printFile <- function(o, n = 1e5) cat(readChar(o, n))
printFile("dotArma.cpp")
```
Notice the `Rcpp::depends(RcppArmadillo)` attribute at the top. To add this function to the `mypack` package (see the previous section), we copy the `.cpp` file to the appropriate directory:
```{r}
system("cp dotArma.cpp mypack/src/dotArma.cpp")
```
We then call `compileAttributes` on the package folder:
```{r}
library(Rcpp)
compileAttributes("mypack")
```
This creates the appropriate `C++` and `R` wrappers in `mypack/src/RcppExports.cpp` and `mypack/R/RcppExports.R`. We are not quite ready to build the package, because `compileAttributes` warns us that our code depends on `RcppArmadillo`, but the `DESCRIPTION` file does not mention `RcppArmadillo` under `LinkingTo`. This is required to link our package to the `RcppArmadillo` header files, and must be done manually:
```{r}
desc <- readLines("mypack/DESCRIPTION")
kk <- which(grepl("LinkingTo", desc))
desc[kk] <- paste0(desc[kk], ", RcppArmadillo")
writeLines(desc, "mypack/DESCRIPTION")

printFile("mypack/DESCRIPTION")
```
We recompile the attributes again: 
```{r}
compileAttributes("mypack")
```
so that `RcppArmadillo` is now included in `RcppExports.cpp`:
```{r, comment=NA}
printFile("mypack/src/RcppExports.cpp", 300)
```
Our package is ready to be built and installed:
```{r, message = FALSE}
system("R CMD build mypack")
system("R CMD INSTALL mypack_1.0.tar.gz")
```
Here we used `RcppArmadillo` as an example, but if you want to depend on a different package (e.g., `RcppGSL`, `RcppMLPACK`, etc), the process is exactly the same.

### Making your package's `C++` code callable from other packages

Having installed the `mypack` package, we can now use its exported `R` function. For instance, we could do:
```{r, eval = FALSE}
# NB do no run this, because we will re-install mypack later in the script, and 
# loading it here would cause problems
library(mypack) 
dotArma(5:1, 5:1)
# 55
```
If we wanted to build another package (say, `secondPack`) that depends on `mypack`, we could of course import `dotArma` from `mypack` by adding the `importFrom(mypack, dotArma)` line to the `NAMESPACE` file of `secondPack`. But what if we wanted to call the `_mypack_dotArma_I` `C++` function directly in the `C++` code contained in `secondPack`? Or what if we wanted to call `_mypack_dotArma_I` in some `R` function containted in `secondPack` by using:
```{r, eval = FALSE}
.Call("_mypack_dotArma_I", 1:5, 1:5, PACKAGE = "mypack")
# 55
```
? This section explains how do this using `Rcpp` attributes.

Strictly speaking, if we created a package `secondPack` that depended on `mypack` (via `Depends` in descriptions file) and that containted a `.Call` to `_mypack_dotArma_I`, the code would work. But if we were to check the package via `R CMD check secondPack` we would get:
```{r, eval = FALSE}
* checking foreign function calls ... NOTE
Foreign function call to a different package:
  .Call("_mypack_dotArma_I", ..., PACKAGE = "mypack")
See chapter ‘System and foreign language interfaces’ in the ‘Writing R
Extensions’ manual.
```
This is because we are calling a `C++` function that has not been exported by `mypack`. Only its `R` wrapper (`dotArma()`) has been exported. The `Rcpp::interfaces` attribute provides a simple way of exporting a `C++` function from a package. In particular, consider the following function: 
```{r, comment=NA}
printFile("dotArma_2.cpp")
```
Here the `// [[Rcpp::interfaces(cpp)]]` makes so that the `dotArma_I` will be exported from the package. To see how this works, we add this file to the source code of `mypack`:  
```{r}
system("cp dotArma_2.cpp mypack/src/dotArma_2.cpp") 
```
we compile the attributes:
```{r}
compileAttributes("mypack")
```
The `mypack/src/RcppExports.cpp` generated by `compileAttributes` is quite long, and we do not print it out here. But the key lines are:
```{r engine = 'Rcpp', eval = FALSE, comment = NA}
// registerCCallable (register entry points for exported C++ functions)
RcppExport SEXP _mypack_RcppExport_registerCCallable() { 
    R_RegisterCCallable("mypack", "_mypack_dotArma2", (DL_FUNC)_mypack_dotArma_I2_try);
    R_RegisterCCallable("mypack", "_mypack_RcppExport_validate", (DL_FUNC)_mypack_RcppExport_validate);
    return R_NilValue;
}
```
This means that the `_mypack_dotArma2` function (a `C++` wrapper around our `dotArma_I` function) has been registered as callable from other packages. To make it more easily accessible from other packages, `compileAttributes` also created the following header files:
```{r}
printFile("mypack/inst/include/mypack.h")
printFile("mypack/inst/include/mypack_RcppExports.h")
```
This looks quite complicated, but the key points that must be understood are that:

   - `mypack.h` is a header file with the same name as our package, and it simply includes 
     the second header file;
   - `mypack_RcppExports.h` defines the `mypack` namespace (again, the name of our package), which
      contains the `arma::vec dotArma2(arma::vec x1, arma::vec x2)` function. This is the only `C++`
      function that we are exporting from the package via the `Rcpp::interfaces(cpp)` attribute.
      The name `dotArma2` is the one we have specified via the `Rcpp::export(dotRcpp2)` attribute.
   - `dotArma2` is the `C++` function that we are going to call from dependent packages, as we will 
      illustrate later. It can be seen as a wrapper around our original `dotArma_I` `C++` function.

This is all that need to be done to be able to call the `dotArma2` `C++` function from other packages. To demonstrate this, we create a new package:
```{r, message = FALSE}
Rcpp.package.skeleton("secondPack")
```
We add `RcppArmadillo` and `mypack` to the `LinkingTo` field of its description file:
```{r}
desc <- readLines("secondPack/DESCRIPTION")
kk <- which(grepl("LinkingTo", desc))
desc[kk] <- paste0(desc[kk], ", RcppArmadillo", ", mypack")
writeLines(desc, "secondPack/DESCRIPTION")
```
Then we add the following function:
```{r, comments = NA}
printFile("secondDot.cpp")
```
to the `secondPack` package:
```{r}
system("cp secondDot.cpp secondPack/src/secondDot.cpp") 
```
Notice that `secondDot.cpp` includes the header `mypack.h`, which provides a definition of the `dotArma2` function, inside the `mypack` namespace. Hence, to use the `dotArma2` function in the new package we are following the same steps that we used to use functions from `RcppArmadillo`, namely:

   - add `mypack` to `LinkingTo` in the `Description` file;
   - add `#include <mypack.h>` to include the package header;
   - use `mypack::` to find `C++` functions in the package namespace.
   
To see whether this works, we compile and install `mypack`:   
```{r, message = FALSE}
system("R CMD build mypack")
system("R CMD INSTALL mypack_1.0.tar.gz")
```
Then we compile the attributes in `secondPack`:
```{r}
compileAttributes("secondPack")
```
and we install it:
```{r, message = FALSE}
system("R CMD build secondPack")
system("R CMD INSTALL secondPack_1.0.tar.gz")
```
Let's see whether we can load it and use the new function:
```{r, message = FALSE}
library(secondPack)
secondDot(1:5, 1:5)

t(1:5) %*% 1:5
```
It works fine! Hence, the `Rcpp::interfaces` attribute allowed us to make the `dotArma_I2` accessible from other packages at `C++` level. There are a few thing to points out:

   - the `Rcpp::interfaces` attribute applies to all the functions that appear in the relevant file, as long as
     they are preceded by the `Rcpp::export` attribute. In our case, the `dotArma_2.cpp` contains only one 
     function that is exported via `Rcpp::export`.
   - the `Rcpp::interfaces(cpp)` attributes make so that a `C++` wrapper for `dotArma_I2` is created and exported
     from the package. The attribute `Rcpp::interfaces(r, cpp)` would lead to the creation of both a `C++` and an 
     `R` wrapper for `dotArma_I2`. That is, the package would export an `R` function called `dotArma2` and a `C++`
     function with the same name would be defined in the `mypack.h` header. `Rcpp::interfaces(r)` does not do 
     anything useful, as the `dotArma` `R` wrapper would be created and exported anyway by the `Rcpp::export`
     attribute.
   - the `mypack` package is now compatible with the `Rcpp::depends` attribute. For example, now we can do:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo, mypack)]]
#include <mypack.h>

using namespace Rcpp;

// [[Rcpp::export(dotSource)]]
arma::vec dotSource_I(arma::vec x1, arma::vec x2)
{
  arma::vec out(1);
  
  out = mypack::dotArma2(x1, x2);

  return out;
}
')

dotSource(3:1, 3:1)
```
Notice that we don't need to include `RcppArmadillo.h`, because it's already included in `mypack.h`, while both `RcppArmadillo` and `mypack` have to appear in `Rcpp::depends`. This is because both packages are needed to correctly set the compilation environment (e.g., `sourceCpp` will use the compilation flags `-I"some_folder/RcppArmadillo/include"` `-I"some_folder/mypack/include"` to link against `RcppArmadillo.h` and `mypack.h`). 


### Using an `R` package to make a `C++` library available to other packages

Suppose that we have a `C++` library, which consists simply of the following header file:
```{r, comment = NA}
printFile("smart/smart.h")
```
This is called a header-only library, because the functions definition and implementation are both found in the header files, which are denoted by `.h` or `.hpp`. In contrasts, standard `C++` libraries use header files to detail the definitions of the main function that will be used by the library's users, while the implementation and the internal functions are contained in `.cpp` files. For our purposes, the advantage of header-only libraries (such as `Armadillo`) is that it is simple to include their code in an `R` package and to make so that code can be used by other `R` packages. 

To do this, we first move the folder containing the library to the the `inst/include` sub vounder of our package:
```{r}
system("cp -a smart mypack/inst/include/smart")
```
We then have to change the `mypack.h` header (which was automatically generated by `compileAttributes`) to:
```{r, comment = NA}
system("cp -a mypack_v2.h mypack/inst/include/mypack.h")
printFile("mypack/inst/include/mypack.h")
```
The main change is that we added `#include "smart/smart.h"` to include the new library. We can now re-build and re-install our package:
```{r}
system("R CMD build mypack")
system("R CMD INSTALL mypack_1.0.tar.gz")
```
And now the `smart` `C++` library is available via our package:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo, mypack)]]
#include <mypack.h>

using namespace Rcpp;

// [[Rcpp::export(smartSquare)]]
NumericVector smartSquare_I(NumericVector x)
{
  NumericVector out(x.length());

  for(int ii = 0; ii < x.length(); ii++)
  {
   out[ii] = smart::mysquare(x[ii]);
  }

  return out;
}
')

smartSquare(1:5)
```
It works! This is pretty much the mechanisms used by `RcppArmadillo` to make the `Armadillo` library available to other `R` packages and to `Rcpp` programs compiled via `sourceCpp` (of course, `RcppArmadillo` does more than that, for instance, is extendes the `Rcpp::wrap` and `Rcpp::as` to facilitate conversion between `Armadillo` and `Rcpp` object classes).

For another simple example on shipping header-only libraries via an `R` package, see [this template](https://github.com/r-pkg-examples/rcpp-shared-cpp-functions). For a more complex example, see the [sitmo](https://github.com/coatless/sitmo) package.

