---
title: 1. Using OpenMP in Rcpp
weight: 1
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

Here we describe some basic examples on the use of OpenMP to parallelise Rcpp code. We assume that the reader has some basic familiarity with OpenMP. If you need a refresher, have a look at [these notes](http://chryswoods.com/beginning_openmp/) from Chris Wood.

### Basic examples

The following `Rcpp` program simply pauses the system for `sec` seconds:
```{r}
library(Rcpp)
sourceCpp(code = '
#include <unistd.h>
#include <Rcpp.h>

// [[Rcpp::export(wait_a_second)]]
bool wait_a_second(int sec)
{
 for(size_t ii = 0; ii < sec; ii++)
 { 
  sleep(1);
 }
 return 1;
}
')

system.time( wait_a_second(2) )[3]
```
Where `sleep` is defined in `unistd.h`. The following function uses OpenMP to wait `sec` seconds on `ncores` in parallel:
```{r}
sourceCpp(code = '
#include <unistd.h>
#include <Rcpp.h>

// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(wait_a_second_omp)]]
bool wait_a_second_omp(int sec, int ncores)
{

 #if defined(_OPENMP)
  #pragma omp parallel num_threads(ncores)
  #pragma omp for
 #endif
 for(size_t ii = 0; ii < sec; ii++)
 { 
  sleep(1);
 }
 
 return 1;

 }
')
```
Noet that we used the `Rcpp::plugins` attribute to include OpenMP in the compilation of the `Rcpp` function. The key OpenMP directives are 
```{r engine='Rcpp', eval = FALSE}
#pragma omp parallel num_threads(ncores)
```
which indicates the beginning of a parallel section, to be executed on `ncores` parallel threads, and
```{r engine='Rcpp', eval = FALSE}
#pragma omp for
```
which tells the compiler that the `for` loop can be run in parallel. Let try if this works:
```{r, cache = TRUE}
system.time(wait_a_second_omp(4, 1))[3]

system.time(wait_a_second_omp(4, 4))[3]

system.time(wait_a_second_omp(16, 16))[3]
```
It seems so! Note that the speed up is linear in the number of threads in this case, because each thread is essentially doing nothing. That is, there is no competition for computing resources (i.e., floating-point processing units). To illustrate that this is generally not the case, let us consider the following `Rcpp` function:
```{r}
sourceCpp(code = '
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export(allFiniteSeq)]]
bool allFiniteCpp(NumericVector x)
{

 size_t n = x.size();
 double out = 0;
 
 for(size_t ii = 0; ii < n; ii++)
 {
  out += x[ii];
 }
 
 return R_FINITE(out);

 }
')
```
which returns `TRUE` is all the elements of `x` are finite, `FALSE` otherwise. We can use OpenMP to parallelise it as follows:
```{r}
sourceCpp(code = '
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(allFiniteOMP)]]
bool allFiniteOMP(NumericVector x, int ncores)
{

 size_t n = x.size();
 double out = 0;
 
 #if defined(_OPENMP)
  #pragma omp parallel for num_threads(ncores)
 #endif
 for(size_t ii = 0; ii < n; ii++)
 {
  out += x[ii];
 }
 
 return R_FINITE(out);

 }
')
```
On my Intel i7-3820 3.60GHz CPU with 4 cores 8 threads I get the following timing: 
```{r, results='hide'}
x <- rnorm(1e7)

library(microbenchmark)
microbenchmark(all(is.finite(x)),
               allFiniteSeq(x),
               allFiniteOMP(x, 1), 
               allFiniteOMP(x, 4),
               allFiniteOMP(x, 8), 
               allFiniteOMP(x, 16),
               unit = 'relative')
```
```{r, eval = FALSE}
Unit: relative
                expr       min        lq       mean    median         uq      max neval
   all(is.finite(x)) 18.892823 17.973095 14.2843653 17.272120 13.0616667 7.876008   100
     allFiniteSeq(x)  4.477392  4.282347  3.2067998  4.076221  2.5890351 1.903284   100
  allFiniteOMP(x, 1)  4.486162  4.288818  3.2039762  4.082022  2.5897148 1.997958   100
  allFiniteOMP(x, 4)  1.198740  1.238068  0.9849882  1.197137  0.7789038 1.127209   100
  allFiniteOMP(x, 8)  1.000000  1.000000  1.0000000  1.000000  1.0000000 1.000000   100
 allFiniteOMP(x, 16)  1.045107  1.103636  1.0457466  1.165450  1.0583197 1.170184   100
```
Note that the `Rcpp` version is around 4 times fast that the `R` code, and that with 4 threads we get roughly a linear speed-up. However, the speed-up gains are negligible after that, and using 16 threads is actually detrimental. This is because a processor with 4 physical cores has probably 4 floating-point (FLOP) units, hence trying to parallelise numerical computations on more than 4 threads does not make sense (the threads will start to compete for the same FLOP units). Note also that above we used:
```{r engine='Rcpp', eval = FALSE}
#pragma omp parallel for num_threads(ncores)
```
which is a shortcut for:
```{r engine='Rcpp', eval = FALSE}
#pragma omp parallel num_threads(ncores)
#pragma omp for
```

One important thing to point out is that R's C API and the Rcpp API are not thread-safe in generally. Hence, doing things such as:
```{r engine='Rcpp', eval = FALSE}
#pragma omp parallel
{
 NumericVector x(10);
 int u = x.length();
}
```
is not safe and might behave unexpectedly. In our code for the `allFiniteOMP` function above we do `out += x[ii]`, which selects elements of the `NumericVector` `x` in parallel. This also not guaranteed to be thread-safe, hence it is preferrable to use the thread-safe data structures provided by the [RcppParallel R package](https://cran.r-project.org/web/packages/RcppParallel/index.html), which will be discussed later in these notes. In the following section we show how to perform parallel randon number generation using OpenMP and a thread-safe RNG. 


### Parallel random number generation using OpenMP 

As other parts of R's C API, R's RNG is not thread safe. Hence `Rcpp` code such as:
```{r engine='Rcpp', eval = FALSE}
#pragma omp parallel
{
x = R::rbeta(a, b);
}
```
will produce invalid results and might crash your R session. Hence, we must adopt a thread-safe RNG, such as the one provided by the [sitmo R package](https://cran.r-project.org/web/packages/sitmo/index.html).

Before explaining how to use `sitmo` in parallel, let us consider the following sequential function:
```{r}
sourceCpp(code = '
#include <Rcpp.h>
#include <sitmo.h>

// [[Rcpp::depends(sitmo)]]

// [[Rcpp::export(sumunif_sitmo)]]
Rcpp::NumericVector sumunif_sitmo(unsigned int n,
                                  unsigned int nstep,
                                  double seed) {
  Rcpp::NumericVector out(n);

  uint32_t coreseed = static_cast<uint32_t>(seed);
  sitmo::prng eng(coreseed);
  
  double mx = sitmo::prng::max();
  double tmp = 0;

   for(unsigned int ii = 0; ii < n; ++ii) {
     tmp = 0.0;
     for(unsigned int kk = 0; kk < nstep; ++kk){
      tmp += eng() / mx;
     }
     out[ii] = tmp;
   }
    
  return out;
}
')
```
This function simply simulates `n` sums of `nstep` uniform U(0, 1) random variables. Hence, it is equivalent to the following R code:
```{r, eval = FALSE}
rowSums(matrix(runif(n*nstep), n, nstep))
```
Let us examine the key steps in `runif_sitmo_omp`:

   - we state that our code depends on the `sitmo` package via the `Rcpp::depends` attribute;
   - we convert the seed provided by R to type `uint32_t` via `uint32_t coreseed = static_cast<uint32_t>(seed);`
   - we set up an RNG via `sitmo::prng eng(coreseed);`
   - having done this, calling `eng()` will generate an uniform random number between 0 and `sitmo::prng::max()`, 
     hence we use `((double) eng()) / mx` to rescale the output of `eng()` to (0, 1).
     
Let's try it:
```{r, fig.align='center', fig.width=6, fig.height=3}
n <- 1e3
nstep <- 1000

par(mfrow = c(1, 2))
hist(rowSums(matrix(runif(n*nstep), n, nstep)), main = "R", xlab = "x")
hist(sumunif_sitmo(n, nstep = nstep, seed = 1), main = "Sitmo", xlab = "x")
```
The histograms show the Central Limit Theorem kicking in. 

Now, let us look at a parallel version:
```{r}
sourceCpp(code = '
#include <Rcpp.h>
#include <sitmo.h>

#ifdef _OPENMP
#include <omp.h>
#endif

// [[Rcpp::depends(sitmo)]]
// [[Rcpp::plugins(openmp)]]

// [[Rcpp::export(sumunif_sitmo_omp)]]
Rcpp::NumericVector sumunif_sitmo_omp(unsigned int n,
                                      unsigned int nstep,
                                      Rcpp::NumericVector seeds) {
  Rcpp::NumericVector out(n);
  
  unsigned int ncores = seeds.size();
  
  #ifdef _OPENMP
  #pragma omp parallel num_threads(ncores)
  {
  #endif
  
   uint32_t coreseed = static_cast<uint32_t>(seeds[0]);
   
   #ifdef _OPENMP
    coreseed = static_cast<uint32_t>(seeds[omp_get_thread_num()]);
   #endif
   
   sitmo::prng eng(coreseed);
   
   double mx = sitmo::prng::max();
   double tmp = 0;
  
   #ifdef _OPENMP
    #pragma omp for 
   #endif
   for(unsigned int ii = 0; ii < n; ++ii) {
     tmp = 0.0;
     for(unsigned int kk = 0; kk < nstep; ++kk){
      tmp += eng() / mx;
     }
     out[ii] = tmp;
   }
    
  #ifdef _OPENMP
  }
  #endif
  
  return out;
}
')
```
Note that this code:
```{r engine='Rcpp', eval = FALSE}
#ifdef _OPENMP
#pragma omp parallel num_threads(ncores)
{
#endif
```
marks the beginning of a parallel section of the code, hence the variables declared in this section (e.g, `coreseed`, `mx`) and the RNG `eng` are private to each parallel thread. This code:
```{r engine='Rcpp', eval = FALSE}
#ifdef _OPENMP
  coreseed = static_cast<uint32_t>(seeds[omp_get_thread_num()]);
#endif
```
provides the seed to each RNG instance using the ID number of each thread (`omp_get_thread_num()` is defined in `omp.h`). The rest of the code should be relatively clear. Let us test it:
```{r, fig.align='center', fig.width=6, fig.height=3}
par(mfrow = c(1, 2))
hist(rowSums(matrix(runif(n*nstep), n, nstep)), main = "R", xlab = "x")
hist(sumunif_sitmo_omp(n, nstep = nstep, seed = 1:4), main = "Sitmo", xlab = "x")
```
It seems to work fine. Let's look at the CPU time:
```{r, results = 'hide'}
microbenchmark(R = rowSums(matrix(runif(n*nstep), n, nstep)),
               sitmo = sumunif_sitmo(n, nstep = nstep, seed = 1),
               sitmo_omp1 = sumunif_sitmo_omp(n, nstep = nstep, seeds = 1),
               sitmo_omp4 = sumunif_sitmo_omp(n, nstep = nstep, seeds = 1:4), 
               sitmo_omp16 = sumunif_sitmo_omp(n, nstep = nstep, seeds = 1:16), 
               unit = "relative", 
               times = 100
               )
```
On the system described above, I get:
```{r, eval = FALSE}
Unit: relative
        expr       min        lq      mean    median        uq       max neval
           R 19.129317 18.220434 15.896634 16.341999 15.275812 22.446736   100
       sitmo  4.895919  4.664486  3.910135  4.147784  3.596815  2.421466   100
  sitmo_omp1  4.887580  4.652173  3.924101  4.135457  3.600384  2.472559   100
  sitmo_omp4  1.231661  1.769258  1.467673  1.584121  1.359834  1.210142   100
 sitmo_omp16  1.000000  1.000000  1.000000  1.000000  1.000000  1.000000   100
```
Hence, `Rcpp` is four times faster than `R`, and we get an acceleration of factor of a little bit more than 2 by using 4 cores. The acceleration levels off if we use more than 16 threads, but in this case we seem to be able to get some extra speed gains by using more threads than actual physical cores. It is difficult to know why this is the case.



