---
title: 1. Using R's C API
weight: 1
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

The purpose of this section is explaining how to interface R with C/C++ using the raw C API provided by R. In practice it is generally preferable to use the `Rcpp` package to interface R with C++, but it is useful to know how R’s C API work to have an idea of what is going on under the hood when you use `Rcpp`. Besides, many of the functions in base R are written in C using the R's C API and do not use `Rcpp`. This section assumes that you have some basic knowledge of C.

### Creating C functions that are callable from R

To provide some motivation, consider the following data:
```{r}
library(gamair)
library(tibble)
data(chicago)
chicago <- as_tibble(chicago)
chicago
```
The `chicago` data set contains data on daily air pollution and death rate for Chicago (see `?chicago` for details). Given this data, one might be interested in analysing the relation between the number of deaths and air pollution. In particular, one could consider the following GAM model (taken from Wood, 2017, page 347)
$$
\text{death}_t \sim \text{Poi}(\lambda_t),
$$
$$
\log \lambda_t = f_1(\text{time}_t) + f_2(\text{pm10}_t) + f_3(\text{so2}_t) + f_4(\text{o3}_t) + f_5(\text{tmpd}_t), 
$$
Which can be fitted using `mgcv` as follows: 
```{r, message = FALSE}
library(mgcViz)
fit0 <- bam(death ~ s(time, k = 200) + s(pm10median) + s(so2median) + s(o3median) +
                    s(tmpd),
            data = chicago, family = poisson, discrete = TRUE)
```
The we can use `mgcViz` to plot the effect of the three pollutants and of the daily temperature:
```{r, fig.align='center', fig.height=5.5, fig.width=7}
fit0 <- getViz(fit0)
print(plot(fit0, select = 2:5), pages = 1)
```
The effect of `so2` doesn't seem to be important, hence it might make sense to remove it from the model. Another thing that we could try, it to check whether the number of deaths depends on the temperatures (`tmpd`) registered in the last few days, rather than only on the temperature on the same day. Rather than simply using lagged values of the temperature, we could consider building a new variable obtained by exponentially smoothing the temperature, that is
$$
\text{tmpdSmooth}_t = \alpha \, \text{tmpdSmooth}_{t-1} + (1 - \alpha) \, \text{tmpd}_t,
$$
with $\alpha \in (0, 1)$.

Now assume that we want to implement a function, say, `expSmooth` for obtaining the exponential smooth of an variable in C, and that we want to be able to call it from R. The C code for such a function is:  
```{r engine='Rcpp', eval = FALSE}
#include <R.h>
#include <Rinternals.h>

SEXP expSmooth(SEXP y, SEXP ys, SEXP n, SEXP a)
{
  int ni;
  double *xy, *xys;
  double ai;
  
  xy = REAL(y); 
  xys = REAL(ys);
  ni = INTEGER(n)[0];
  ai = REAL(a)[0];

  xys[0] = xy[0];
  for(int i = 1; i < ni; i++){
    xys[i] = ai * xys[i-1] + (1 - ai) * xy[i];
  }
  
  return R_NilValue;
}
```
Before explaing how `expSmooth` can be called from R, let's try to break it down. The first two lines are there to include two header files that contain some R-related C functions and macros. For example `REAL` is defined in `Rinternals.h` as can be verified by looking at the source code by doing (not shown here):
```{r, eval = FALSE}
rinternals <- file.path(R.home("include"), "Rinternals.h")
file.show(rinternals)
```
The `expSmooth` function returns an `SEXP`, which is also the class of its inputs. A object of class `SEXP` is a pointer to an S expression, that is a pointer on an object in the R environment. Hence, importantly, the inputs are passed to `expSmooth` by reference and are not copied. Here our inputs are:
  
   - `y`, the vector to be smoothed;
   - `ys`, the vector which will contain the smoothed version of `y`;
   - `n`, a scalar indicating the length of `y` and `ys`;
   - `a`, the $\alpha$ coefficient.

Even though, all the arguments of `expSmooth` are of class `SEXP`, is important to know that `SEXP` is a variant type, that is a C object whose subtype is known only at runtime, with subtypes that cover all data structures in R. Examples of subtypes are:
   
   - `REALSXP` a numeric vector such as `c(1.2, 0.45)`;
   - `LGLSXP` a logical vector such as `c(TRUE, FALSE, FALSE)`;
   - `INTSXP` an integer vector such as `c(2L, 34L, 1L)`;
   - `VECSXP` a list such as `list("a" = 2, "b" = c(5, 4))`.
   
See the [R internal documentation](https://cran.r-project.org/doc/manuals/R-ints.html#R-Internal-Structures) for a complete list.
   
The first lines of the function's body are:
```{r engine='Rcpp', eval = FALSE}
int ni;
double *xy, *xys;
double ai;
```
where we are creating four C objects, which are the initialized using:
```{r engine='Rcpp', eval = FALSE}
xy = REAL(y); 
xys = REAL(ys);
ni = INTEGER(n)[0];
ai = REAL(a)[0];
```
We are using the [macro function](https://gcc.gnu.org/onlinedocs/cpp/Macro-Arguments.html) `REAL` to access object `y`. In particular, `REAL` returns a double pointer to the real part of the `SEXP` `y`. Simularly, `INTEGER(n)` returnes a pointer to the integer part of `n` and by doing `[0]` we are extracting the value of its first element. Hence, here there is the assumption that the subtype of `y` is `REALSXP` and that the subtype of `n` is `INTSXP`. Otherwise if, for example, `n` was a real vector of type `REALSXP`, calling `INTEGER` would trigger an error at runtime, as we shall see.

Then computation actually happens here:
```{r engine='Rcpp', eval = FALSE}
xys[0] = xy[0];
for(int i = 1; i < ni; i++){
  xys[i] = ai * xys[i-1] + (1 - ai) * xy[i];
}
```
In the first line we are initializing by setting $\text{tmpdSmooth}_1 = \text{tmpd}_1$, than the rest is straighforward C code. Notice that the function returns:
```{r engine='Rcpp', eval = FALSE}
return R_NilValue;
```
where is the C equivalent of R's `NULL`. Hence the function is not returning anything useful, but the smoothed variable is obtained by modifying the $xys$ vector in place (that is, without making a copy).

### Calling expSmooth from R

Now, assume that `expSmooth` is contained in a file called `expSmooth.c`, which in our case can be found in our working directory, as you can see from the following call to the command line: 
```{r}
system("ls *.c")
```
Here we will explaing how to call `expSmooth` using the `.Call` interface in R. `.Call` can be use to call C functions that have accept variable of class `SEXP` as inputs and return and object of class `SEXP`. Indeed, we created the `expSmooth` function with this requirement in mind.

The first thing that we need to do is to compile the C code as follows
```{r}
system("R CMD SHLIB expSmooth.c")
```
This will create two files, `expSmooth.o` and `expSmooth.so`, containing binary code. The `.so` object is a share object which can be loaded in R by doing: 
```{r}
dyn.load("expSmooth.so")
```
Having done this, the `expSmooth` function is now available at R level:
```{r}
is.loaded("expSmooth")
```
See the [Creating shared objects](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#Creating-shared-objects) section of the [Writing R Extensions](https://cran.r-project.org/doc/manuals/r-release/R-exts.html) manual for more details on how this works.

Having loaded `expSmooth` in R, we can now call it using `.Call`:
```{r}
nch <- nrow( chicago )
tmpSmooth <- numeric(nch)

.Call("expSmooth", chicago$tmpd, tmpSmooth, nch, 0.9)
```
As expected the function returns 'NULL', but the smoothed temperature is now contained in `tmpSmooth`:
```{r, fig.align='center', fig.height=4, fig.width=5.5}
plot(chicago$tmpd[1:1000], col = "grey", ylab = "Temp")
lines(tmpSmooth[1:1000], col = 2)
```
To check whether our function actually works, let's compare it with an R version of it:
```{r} 
expSmoothR <- function(x, a){

  n <- length(x)
  xs <- x

  for(ii in 2:n){
   xs[ii] <- a * xs[ii - 1] + (1 - a) * x[ii]
  }

  return(xs)

}
```
The output of the R and C version is identical:
```{r}
max( abs(tmpSmooth - expSmoothR(chicago$tmpd, 0.9)) )
```

Having verifying that our C function for exponential smoothing does the right thing, let's see how it compares with its R version in terms of computing time: 
```{r}
smoR <- function() expSmoothR(chicago$tmpd, 0.9)
smoC <- function() .Call("expSmooth", chicago$tmpd, tmpSmooth, nch, 0.9)

library(microbenchmark)
microbenchmark(smoR(), smoC(), times = 500)
```
So the C version is more than 10 times faster. The comparison it somewhat unfair, because the R function is allocating memory containing the smoothed temperature at every call, while the C version is re-using the memory allocated to the `tmpSmooth` vector. However, the ability to overwrite R objects without copying them is one of the reasons for interfacing R with C. 

We now go back to our original application, and fit a model which excludes `so2` with didn't seem to matter and that includes the effect of smoothed temperature:
```{r, message = FALSE}
chicago$tmpSmooth <- tmpSmooth
fit1 <- bam(death ~ s(time, k = 200) + s(pm10median) + s(o3median) +
                    s(tmpd) + s(tmpSmooth),
                   data = chicago, family = poisson, discrete = TRUE)
```
We now plot the effects of the two temperature effects:
```{r, fig.align='center', fig.height=3.5, fig.width=7}
fit1 <- getViz(fit1)
print(plot(fit1, select = 4:5), pages = 1)
```
The plots shows that the effect of smoothed temperature is quite important, and the new model seems to do better than the old model in terms of AIC:
```{r}
AIC(fit0, fit1)
```

### Creating R objects within C code

Here we try to refine the `expSmooth` function created above. Firstly notice that, while `expSmooth` does not make any explicit assumption about the subtype of its `SEXP` input, trying to smooth the number of death leads to:
```{r, eval = FALSE}
.Call("expSmooth", chicago$death, tmpSmooth, nch, 0.9)
# Error: REAL() can only be applied to a 'numeric', not a 'integer'
```
Because `xy = REAL(y);` triggers an error which signal that `y` should be a vector of reals, not of integers. To smooth the number of death we must pass `as.double(chicago$death)`. However, `expSmooth` does not check for other things, for example there nothing guaranteeing that argument `n` actually matches the length of the input vectors `y` and `ys`. For example, the following code:
```{r, eval = FALSE}
.Call("expSmooth", chicago$tmpd, tmpSmooth, nch + 10L, 0.9) # do NOT run this
```
might run without any indication that something wrong has happened. But $n$ is 10 units longer than `length(chicago$tmpd)`, which mean that the last 10 iterations of this loop:
```{r engine='Rcpp', eval = FALSE}
for(int i = 1; i < ni; i++){
  xys[i] = ai * xys[i-1] + (1 - ai) * xy[i];
}
```
are reading outside the memory allocated to `xy` and writing outside the memory allocated to `xys`. This is a serious problem, and our current version of `expSmooth` is entirely relying on the assumption that the user will call it with the right arguments. 

A safer version of `expSmooth` might be something like this:
```{r engine='Rcpp', eval = FALSE}
#include <R.h>
#include <Rinternals.h>

SEXP expSmooth2(SEXP y, SEXP a)
{
  int ni;
  double *xy, *xys;
  double ai;
  SEXP ys;
  
  y = PROTECT(coerceVector(y, REALSXP));
  ni = length(y);
  ys = PROTECT(allocVector(REALSXP, ni));
  
  ai = REAL(a)[0];
  xy = REAL(y); 
  xys = REAL(ys);
  
  xys[0] = xy[0];
  for(int i = 1; i < ni; i++){
    xys[i] = ai * xys[i-1] + (1 - ai) * xy[i];
  }
  
  UNPROTECT(2);
  
  return ys;
}
```
Notice that `expSmooth2` takes only two functions: the variable to be smoothed `y` and the smoothing coefficient `a`. Then here:
```{r engine='Rcpp', eval = FALSE}
y = PROTECT(coerceVector(y, REALSXP));
ni = length(y);
```
We are using `coerceVector` to convert `y` to subtype `REALSXP` hence this analogous to doing `as.double` in R. The output of `coerceVector` might be protected from R's garbage collector using `PROTECT`, otherwise the memory allocated to `y` might be deallocated before we finished using it. All R objects allocated in C code must be protected, as failing to do so might lead to intermittent crashes and hard to find bugs. See this [blog post](https://developer.r-project.org/Blog/public/2019/04/18/common-protect-errors/) for more details. The next line just extracts the length of `y` and stores it in `ni`, thus removing the need to pass the length of `y` as an argument.

Then we allocate memory for the smoothed output:
```{r engine='Rcpp', eval = FALSE}
ys = PROTECT(allocVector(REALSXP, ni));
```
where we are allocating memory for a vector of doubles with `ni` elements. This also must be protected, because we are allocated an R object (`allocVector` is defined in `Rinternals.h`). The main loop is the same, but `expSmooth2` ends with:
```{r engine='Rcpp', eval = FALSE}
UNPROTECT(2);
return ys;
```
where we are unprotecting two objects and then returning `ys`. It is important to UNPROTECT as many objects as you have protected, and to UNPROTECT just before returning to R. Failing to UNPROTECT will lead to memory leakage, that is to R objects hanging around in memory while not being useful or accessible in any way. 

Having created our new C function, which can be found in `expSmooth2.c`, we compile it and load it: 
```{r}
system("R CMD SHLIB expSmooth2.c")
dyn.load("expSmooth2.so")
```
We compare its output with that of the R version:
```{r}
max(abs(.Call("expSmooth2", chicago$tmpd, 0.9) - expSmoothR(chicago$tmpd, 0.9)))
```
which shows that it is working correctly. The computing time is also comparable to that of the old C function:
```{r}
smoR <- function() expSmoothR(chicago$tmpd, 0.9)
smoC <- function() .Call("expSmooth", chicago$tmpd, tmpSmooth, nch, 0.9)
smoC2 <- function() .Call("expSmooth2", chicago$tmpd, 0.9)

library(microbenchmark)
microbenchmark(smoR(), smoC(), smoC2(), times = 500)
```
despite the fact that is it allocating memory for the output vector at every call. Importanly, `expSmooth2` is safer and works directly with integer input vectors:
```{r, fig.align='center', fig.height=4, fig.width=5.5}
plot(chicago$death[1:1000], col = "grey")
lines(.Call("expSmooth2", chicago$death, 0.9)[1:1000], col = 2)

```


