---
title: 3. Rcpp basics
weight: 3
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

### Rcpp via R CMD SHLIB: the old but instructive way

To illustrate how `Rcpp` works, let us consider again our C function for exponential smoothing, that is:
```{r engine='Rcpp', eval = FALSE}
#include <R.h>
#include <Rinternals.h>

SEXP expSmooth2(SEXP y, SEXP a)
{
  int ni;
  double *xy, *xys;
  double ai;
  SEXP ys;
  
  y = PROTECT(coerceVector(y, REALSXP));
  ni = length(y);
  ys = PROTECT(allocVector(REALSXP, ni));
  
  ai = REAL(a)[0];
  xy = REAL(y); 
  xys = REAL(ys);
  
  xys[0] = xy[0];
  for(int i = 1; i < ni; i++){
    xys[i] = ai * xys[i-1] + (1 - ai) * xy[i];
  }
  
  UNPROTECT(2);
  
  return ys;
}
```
Now, the same function can be implemented in "raw" Rcpp as follows (as we will explain later this is NOT the recommended way of using `Rcpp`):
```{r engine='Rcpp', eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

RcppExport SEXP expSmoothRcpp_manual(SEXP ySEXP, SEXP aSEXP)
{
  const NumericVector y = as<const NumericVector>(ySEXP);
  const double a = as<const double>(aSEXP);
  
  int ni = y.size();
  NumericVector ys(ni);
  
  ys[0] = y[0];
  for(int i = 1; i < ni; i++){
    ys[i] = a * ys[i-1] + (1 - a) * y[i];
  }
  
  return ys;
}
```
Don't worry if you don't understand what is going on in the code above, we'll explain it below. As you can see the `Rcpp` function for exponential smoothing is much more compact. On the Linux command line it can be compiled by doing:
```{r}
library(Rcpp)
system("export PKG_CXXFLAGS=`Rscript -e \"Rcpp:::CxxFlags()\"` && 
        R CMD SHLIB expSmoothRcpp_manual.cpp")
```
Note that the call to `R CMD SHLIB` is preceded by some extra code, which is needed to let `SHLIB` know where `Rcpp.h` can be found. We then load the compiled code as usual:
```{r}
dyn.load("expSmoothRcpp_manual.so")
```
Let's see whether the new function actually works:
```{r}
system("R CMD SHLIB expSmooth2.c")
dyn.load("expSmooth2.so")

y <- rnorm(100)
max( abs(.Call("expSmoothRcpp_manual", y, 0.9) - .Call("expSmooth2", y, 0.9)) )
```
So the `C` and `Rcpp` functions produce exactly the same output.

Now, let us try to understand how the `Rcpp` function for exponential smoothing works:

   - in the first two lines we are including the `Rcpp.h` header and we are adopting the corresponding
     namespace.
   - `expSmoothRcpp_manual` has `SEXP` inputs and outputs, hence it can be called via `.Call`.
   - `RcppExport` is an alias for `extern "C"`, the reason why we needed the latter is not important here, 
      but it is there to tell the C++ compiler not to mangle the name of the function, otherwise `.Call` 
      would not be able to find it.
   - `const NumericVector y = as<const NumericVector>(ySEXP);` converts the `ySEXP` argument, which has 
      class `SEXP`, to a constant object, `y`, of class `NumericVector`, which simply represents 
      vectors of doubles. The `NumericVector` class and the templated function `as` belong to the `Rcpp` namespace.
      Importantly, `ySEXP` is not copied, and modifying `y` would modify `ySEXP`. We avoid this risk by using `const`.
   - `const double a = as<const double>(aSEXP);` converts `aSEXP` similarly.
   - `int ni = y.size();` here we are using the `size` member function provided by the `NumericVector` class to extract       the size (or length) of `y` and store it into an integer `ni`.
   - `NumericVector ys(ni);` creates a new `NumericVector` `ys` of size `ni`.
   - Then the `for` loop is very similar to the one in `expSmooth2` and we return the smoothed vector `ys`, without
     calling `UNPROTECT`.
     
It is worth repeating that compiling `Rcpp` code via `R CMD SHLIB` is not recommended, and the next section will illustrate some better methods. The main advantage of `Rcpp` illustrated so far is that we didn't explicitly create any object of class `SEXP`, which means that we didn't have to call `PROTECT` and `UNPROTECT` to do memory management. In fact, this was does automatically for us via the `NumericVector` class, which wraps an `R` object of class `SEXP` and protects it from `R`'s garbage collector while it is in scope (that is until we exit the part of the program where the
variable is accessible). Hence, our code is shorter and one source of mistakes (i.e., incorrect memory management) has been removed. `Rcpp` provides many other wrappers around `R` objects, such as:

   - `NumericMatrix` wraps a `numeric` matrix in `R`;
   - `IntegerVector`, `IntegerMatrix` wrap `integer` vectors/matrices in `R`;
   - `List` wraps a `list` in `R`;
   - `Function` wraps a `function` in `R`;
   - see [here](https://thecoatlessprofessor.com/programming/cpp/unofficial-rcpp-api-documentation/#vmld)
     for more classes.
     
The pattern is always the same, upon passing from `R` to a `C++` function, wrap all the `SEXP` inputs via one of the wrapper classes defined in `Rcpp`, so that you will be able manipulate them in `C++` without thinking about `R` memory management. As we will see in the next section, the conversion from `SEXP` to an `Rcpp` wrapper can be handled automatically by `Rcpp`.

### Rcpp via `sourceCpp()`

Consider the following `C++` function:
```{r engine='Rcpp', eval = FALSE}
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export(name = "expSmoothRcpp")]]
NumericVector expSmoothRcpp_I(const NumericVector y, const double a)
{
  int ni = y.size();
  NumericVector ys(ni);
  
  ys[0] = y[0];
  for(int i = 1; i < ni; i++){
    ys[i] = a * ys[i-1] + (1 - a) * y[i];
  }
  
  return ys;
}
```
This is very similar to `expSmoothRcpp_manual` (see above). In fact, if we leave aside the comment:
```{r engine='Rcpp', eval = FALSE}
// [[Rcpp::export(name = "expSmoothRcpp")]]
```
for a moment, the main difference is that now the function signature has changed from:
```{r engine='Rcpp', eval = FALSE}
RcppExport SEXP expSmoothRcpp_manual(SEXP ySEXP, SEXP aSEXP)
```
to:
```{r engine='Rcpp', eval = FALSE}
NumericVector expSmoothRcpp_I(const NumericVector y, const double a)
```
Hence, the function is not using `RcppExport` and its inputs and output are not of class `SEXP`, which means that this function cannot be directly called from `R` using `.Call()`. To make this function accessible from `R`, we can do:
```{r}
sourceCpp("expSmoothRcpp_I.cpp")
```
which compiles `expSmoothRcpp_I`, loads the corresponding dynamic library in `R` (using `dyn.load`) and creates and `R` wrapper called `expSmoothRcpp`. Hence, to test if this worked we do:
```{r}
max( abs(expSmoothRcpp(y, 0.9) - .Call("expSmooth2", y, 0.9)) )
```
Perfect agreement.

Now, to understanding how this works, note that:
```{r}
expSmoothRcpp
```
shows that `expSmoothRcpp` is a simple wrapper around a `.Call` to a C++ function called `<pointer: 0x7ff6985a0520>` (or similar). The R function and the C++ function called in the internal call to `.Call` have been generated by the call to `sourceCpp`. We can get more details about what `sourceCpp` is doing, by using the `verbose` argument:
```{r}
sourceCpp("expSmoothRcpp_I.cpp", rebuild = TRUE, verbose = TRUE)
```
The first part of the text output states that some `extern "C"` functions have been generated. In particular, we see the definition of the function:
```{r engine='Rcpp', eval = FALSE}
RcppExport SEXP sourceCpp_1_expSmoothRcpp_I(SEXP ySEXP, SEXP aSEXP)
```
The fact that the definition of `sourceCpp_1_expSmoothRcpp` is preceded by `RcppExport` and that the inputs and output are of class `SEXP` means that the function can be called from R using `.Call` (remember that this is a requirement of `.Call`). The fact that `sourceCpp_1_expSmoothRcpp` is compatible with `.Call` and that it contains a call to `expSmoothRcpp_I` (our C++ code, above) shows that `sourceCpp_1_expSmoothRcpp` is a wrapper around `expSmoothRcpp_I`, making it accessible from `R`. We now examine the function body:

   - `Rcpp::RObject rcpp_result_gen;` creates an object of class `RObject`, which is defined in `Rcpp.h`. 
      The `RObject` class is a wrapper around an `SEXP` object, which is not copied. Here the object is 
      initially empty. The memory of `RObject` is managed automatically, that is the object is protected
      from garbage collector while it is in scope. The `RObject` is a base class, from which many `Rcpp` classes 
      (such as `NumericVector`, `List`, `Function`, etc) are derived.
   - `Rcpp::RNGScope rcpp_rngScope_gen;` has to do with the initialization of the random number generator,
      which is not important here, but we'll say more about it later.
   - `Rcpp::traits::input_parameter< const NumericVector >::type y(ySEXP);` creates an object `y`, which is a 
      wrapper around `ySEXP`. `y` is a constant `NumericVector`, hence it can be passed to our
      `expSmoothRcpp` function. A similar thing is done for `a`, in the next line of code. Explaining
      how `input_parameter` works is quite involved, but for our purposes this code is equivalent to doing 
      
      `const NumericVector y = as<const NumericVector>(ySEXP);`
      
      As in our original `expSmoothRcpp_manual` function. 
      
So, to make our `expSmoothRcpp_I` function accessible from `R`, `sourceCpp` has written a `C++` wrapper which is compatible with `.Call` and does all the necessary conversions from `SEXP` to `NumericVector`, `double` and potentially many other `Rcpp` or `C++` classes for us. The next lines in the text output show that `sourceCpp` has:

  - dynamically loaded the shared object generated by compiling the `C++` code in `expSmoothRcpp_I.cpp` 
    (see the last line of the text output);
  - built the `expSmoothRcpp` wrapper function.
  
The last thing that needs to be explained is the purpose of the comment preceding our `C++` function definition, that is:
```{r engine='Rcpp', eval = FALSE}
// [[Rcpp::export(name = "expSmoothRcpp")]]
NumericVector expSmoothRcpp_I(const NumericVector y, const double a)
```
Here `Rcpp::exportRcpp` is an example on the Rcpp export attribute, used to declare that the `expSmoothRcpp_I` `C++` function should be callable from R via the automatically generated `R` and `C++` code we have described above. The optional `(name = "expSmoothRcpp")` argument is used to specify the name of the R function that should be used to access the `C++` code. In fact, simply doing:
```{r engine='Rcpp', eval = FALSE}
// [[Rcpp::export]]
NumericVector expSmoothRcpp_I(const NumericVector y, const double a)
```
would have created an `R` wrapper called `expSmoothRcpp_I`. For more on attributes, see the [Rcpp vignette on attributes](http://dirk.eddelbuettel.com/code/rcpp/Rcpp-attributes.pdf), which also clarifies what are the requirements that must met by a `C++` function to be exportable to `R` via `Rcpp::exportRcpp`. 

### Inline Rcpp via `cppFunction()` and `evalCpp()`

While it is generally preferable to have `C++` code stored in its own `.cpp` source files, `Rcpp` provides mechanisms for defining and executing `C++` functions within an R script. In particular, consider the following standard `Rcpp` example:
```{r}
sourceCpp(code = '
#include <Rcpp.h>
using namespace Rcpp;

// [[Rcpp::export(name = "fiboInline1")]]
int fibonacci(const int n) {
 if (n < 2)
  return n;
 else
  return (fibonacci(n-1)) + fibonacci(n-2);
}')
```
which creates a `C++` function for computing the $n$-th element of the Fibonacci sequence:
```{r}
fiboInline1(6)
```
This is equivalent to having defined the function in, say, `fibo.cpp` and having sourced it using `sourceCpp(fibo.cpp)`. An even more compact approach involves using the `cppFunction` function:
```{r}
cppFunction('
int fiboInline2(const int x) {
 if (x < 2)
  return x;
 else
  return (fiboInline2(x-1)) + fiboInline2(x-2);
}
')

fiboInline2(6)
```
Using `Rcpp` in an inline fashion can be convenient we dealing with very small chunks of `C++` code. Simple `C++` expression can be evaluated via `evalCpp`, for instance:
```{r}
evalCpp('1.0 + 1.0')
```
As for `sourceCpp`, we can use the `verbose` argument to look at the code automatically generated by `Rcpp`:
```{r}
evalCpp('1.0 + 1.0', verbose = TRUE)
```
Notice the use of `Rcpp::wrap` when the `get_value` function returns. `wrap` allows to transfer objects from C++ to R, for example here it converts the double 2.0 to a numeric vector in R. An important point is that, when converting from R to C++ using `Rcpp::as<T>`, we need to specify the C++ type via the `T` argument, while the conversion from C++ to R using `wrap` is handled automatically.

### References

- Allaire, J.J., Eddelbuettel, D. and François, R., 2018. Rcpp Attributes. Vignette included in R package Rcpp, URL http://CRAN. R-Project. org/package= Rcpp.
- Eddelbuettel, D., 2013. Seamless R and C++ integration with Rcpp. New York: Springer.
- Eddelbuettel, D. and Balamuta, J.J., 2018. Extending R with C++: A Brief Introduction to Rcpp. The American Statistician, 72(1), pp.28-36.
- Wickham, H., 2014. Advanced r. Chapman and Hall/CRC.

