---
title: 4. Exercises on Rcpp
weight: 4
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

### Computing the c.d.f. of the Tweedie distribution

Here we consider the [Tweedie distribution](https://en.wikipedia.org/wiki/Tweedie_distribution) which includes continuous distributions, such as the normal and gamma, and discrete distributions, such as the Poisson. Its density is 
$$
p(y|\mu,\phi,p)=a(y,\phi,p)\exp\bigg[\frac{1}{\phi}\{y\theta-\kappa(\theta)\}\bigg],
$$
where 
$$
\theta=\frac{\mu^{1-p}}{1-p}\;\;\text{for}\;p\neq1\;\;\;\text{and}\;\;\;\theta=\log\mu\;\;\text{for}\;p=1,
$$
and
$$
\kappa(\theta)=\frac{\mu^{2-p}}{2-p}\;\;\text{for}\;p\neq2\;\;\;\text{and}\;\;\;\theta=\log\mu\;\;\text{for}\;p=2.
$$
with $\mu>0$ being the mean, $\phi>0$ the scale and $1\leq p\leq2$ is such that $\text{var}(y)=\mu^{p}$. As explained in [Dunn and Smyth (2005)](http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.435.5387&rep=rep1&type=pdf), evaluating the Tweedie density requires approximating the factor $a(y,\phi,p)$, which does not have a closed-form expression, using specifically designed numerical methods. Here we consider the simpler problem of numerically approximating the Tweedie c.d.f. $P(y|\mu,\phi,p)$, under the assumption that $a(y,\phi,p)$ is unknown. 

For $1\leq p\leq2$, a Tweedie variable $y$ can be written as the sum of $N$ i.i.d. Gamma distributed random variables $z_{1},\dots,z_{N}$ with shape $-\alpha$ and scale $\gamma$ (hence with mean $-\alpha\gamma$ and variance $-\alpha\gamma^{2}$), while $N$ follows a Poisson distribution with rate $\lambda$. In terms of the original Tweedie parameters, we have
$$
\lambda=\frac{\mu^{2-p}}{\phi(2-p)}, \;\;\; \alpha=(2-p)/(1-p) \;\;\; \text{and} \;\;\; \gamma=\phi(p-1)\mu^{p-1}.
$$
So, if $p_{G}$ indicates the c.d.f. of a Gamma distribution with parameters $-k\alpha$ and $\gamma$, $p_{p}$ is the p.m.f. of a Poisson with rate $\lambda$ and we indicate with $p(Y < y)$ the Tweedie c.d.f., we have
\[
p(Y<y)=p\Big(\sum_{i=1}^{N}z_{i}<y\Big)=\sum_{k=1}^{\infty}p_{G}\Big(\sum_{i=1}^{k}z_{i}<y\Big)p_{P}(N=k),
\]
where the second equality holds due to the Law of Total Probability. Now, we want to avoid computing that infinite sum, hence we approximate it by
$$
p(Y<y)\approx\hat{p}(Y<y)=\sum_{k=k_{min}}^{k_{max}}p_{G}\Big (\sum_{i=1}^{k}z_{i}<y \Big )p_{P}(N=k),
$$
for some $k_{min}$ and $k_{max}$. 

How to choose $k_{min}$ and $k_{max}$? Given that $p_{P}$ is maximal at $k = \text{floor}(\lambda)$, and then monotonically decreases as we move $k$ away from the mode, a reasonable approach is to choose $k_{min}$ and $k_{max}$ such that $p_{P}(N=k) < \epsilon p_{P}\{N=\text{floor}(\lambda)\}$ for $k<k_{min}$ or $k>k_{max}$, for some small $\epsilon$. If we do so, it is clear that the a very pessimistic bound on the absolute approximation error is
$$
|p(Y<y)-\hat{p}(Y<y)| < p_{P}(N<k_{min} \; \text{or} \; N>k_{max}) = 1 - p_{P}(k_{min}<N<k_{max}), 
$$
which is easy to compute. Of course, $k_{min}$ and $k_{max}$ are not known in advance, but we can initialize $k$ to $\text{floor}(\lambda)$ and then increase $k$ until we get to $p_{P}(N=k) < \epsilon p_{P}\{N=\text{floor}(\lambda)\}$, which means that we have found $k_{max}$. Then we can set $k$ to the Poisson mode and decrease $k$ until we find $k_{min}$. All this is implemented in the following R function:
```{r}
pTweedR0 <- function(y, mu, phi, p, eps = 1e-17, log = FALSE){
  
  # Get param lambda, gamma and alpha to be used in gamma and poisson distrib
  la <- mu^(2-p) / ( phi * (2-p) )
  ga <- phi * (p-1) * mu^( p-1 )
  al <- (2-p) / ( 1-p )
  
  # Mode of Poisson is at k = floor(lambda). 
  # If mode is 0, we start from 1 instead.
  k0 <- max(floor(la), 1)
  
  # Poisson density at its mode
  mxlpP <-  dpois(k0, la)
  
  # Get probability contribution at Poisson mode
  pTw <- mxlpP * pgamma(y, shape = - k0 * al, scale = ga)
  
  # Initialize k at mode
  k <- k0 
  lP <- mxlpP
  
  # Sum from mode k = floor(lambda) upward until we find kmax
  while ( lP > mxlpP * eps ){
    k <- k + 1
    lP <- dpois(k, la)
    pTw <- pTw + lP * pgamma(y, shape = - k * al, scale = ga)
  }
  kmax <- k
  
  # Reset k to the Poisson mode, and now go down until we find kmin
  k <- k0
  lP <- mxlpP
  while ( lP > mxlpP * eps && k > 0  ){
    k <- k - 1
    lP <- dpois(k, la)
    pTw <- pTw + lP * pgamma(y, shape = - k * al, scale = ga)
  }
  kmin <- k
  
  if( log ) { pTw <- log( pTw ) }
  
  return( pTw )
  
}
```
Having defined a function for approximating the Tweedie c.d.f., we need to verify whether it is correct. Given that the true c.d.f. is unknown, we compare our function with the `ptweedie` function from the `tweedie` package. Below, we simulate a random vector of Tweedie parameters, and then we compare the (log) c.d.f.s produced by `pTweedR0` with those obtained using `ptweedie`:
```{r, fig.align='center', fig.height=4, fig.width=7}
library("tweedie")
nsim <- 1e3
mu <- runif(nsim, 0, 10)
phi <- 0.01 + rexp(nsim, 1)
p <- runif(nsim, 1.001, 1.999)

lpr1 <- lpr2 <- rep(NA, nsim)
for(ii in 1:nsim){
  y <- rtweedie(n = 1, mu = mu[ii], phi = phi[ii], power = p[ii])
  lpr1[ii] <- pTweedR0(y = y, mu = mu[ii], phi = phi[ii], p = p[ii], log = TRUE)
  lpr2[ii] <- log(ptweedie(q = y, mu = mu[ii], phi = phi[ii], power = p[ii]))
}

par(mfrow = c(1, 2))
plot(lpr1, lpr2, xlab = "pTweedR0", ylab = "ptweedie") 
abline(0, 1)
plot(lpr1 - lpr2, ylab = "pTweedR0 - ptweedie") 
abline(h = 0)
```
The log c.d.f.s are very close, why is not surprising as `ptweedie` uses pretty much the same approximation we are using. 

**Q1 start**: The `pTweedR0` function requires explicitly looping in R, because $k_{min}$ and $k_{max}$ are not known in advance but must be determined iteratively. To address this, create an `Rcpp` version of `pTweedR0` and compare its computational performance with that of `pTweedR0` **Q1 end**.

**Q2 start**: Most the function provided by `R` for evaluating the c.d.f. of a random variable (e.g. `pnorm()`, `pexp()`) are vectorized in all their main arguments (e.g. `pnorm()` is vectorized in the arguments `q`, `mean` and `sd`). Our `pTweedR0` function cannot be not vectorized directly, because $k_{min}$ and $k_{max}$ depend on $y$ and on the Tweedie parameters. Use `Rcpp` to create a vectorized C++ version of `pTweedR0` (that is a version of `pTweedR0` with accepts vector inputs for `y`, `mu`, `phi` and `p`) and compare its performance with that of simply using `pTweedR0` within a `for()` loop in `R` **Q2 end**.







