---
title: 2. RcppArmadillo
weight: 2
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>


### Introduction

Many standard statistical models/algorithms (e.g., linear regression, principal component analysis, ...) require using numerical linear algebra routines, hence this section explains how to perform such computations efficiently using the `RcppArmadillo` `R` package. `RcppArmadillo` provides an interface to the `Armadillo` `C++` numerical linear algebra library, which is highly useful in practice because it provides an interesting balance between performance and ease of use. The syntax of `Armadillo` resembles that of `Matlab`, which is not too different from `R`'s syntax. Part of the performance speed-ups of `Armadillo`, relative to base `R`, are obtained via automatic pooling of several linear algebra operations into one at compile-time.   

### Motivating example 1: matrix-matrix-vector product

To demostrate the capabilities of `Armadillo`, we start by considering a simple example on matrix multiplication. In particular, consider computing the product ${\bf A} {\bf B} {\bf y}$ where ${\bf A}$ and ${\bf B}$ are $d \times d$ matrices and $y$ is a $d$-dimensional vector. In `R` we would compute the product simply by doing `A%*%B%*%y`. Could we get some speed-ups by using `Rcpp`? To verify this, we **try** to create a function that computes the matrix-matrix-vector product in `Rcpp`:
```{r}
library(Rcpp)
sourceCpp(code = '
 #include <Rcpp.h>
 using namespace Rcpp;

 // [[Rcpp::export(name = "MMv_wrong")]]
 NumericVector MMv_I(NumericMatrix A, NumericMatrix B, NumericVector y) {
   return A * B * y;
 }'
)
```
The problem with this function is that it is not equivalent to `A%*%B%*%y`. In fact, it does not computes `A%*%B%*%y`, but something quite different:
```{r}
d <- 3
A <- matrix(1, d, d)
B <- matrix(1, d, d)
y <- 1:d

MMv_wrong(A, B, y)
```
In particular, `A * B` in `MMv_I` is equivalent to `as.vector(A * B)` is `R`, that is it computes the element-wise product of two matrices and then converts the output to a vector. Then one might expect that `A * B * y` in `MMv_I` should be equivalent to:
```{r}
as.vector(A * B) * y
```
in `R`, but this is also wrong because, when multiplying vectors of different length, `Rcpp` does not perform the recycling as `R` does. For example:
```{r}
addDiffLengh <- cppFunction("
 NumericVector addDiffLengh(NumericVector x1, NumericVector x2) {
     return x1 + x2;
}")
addDiffLengh(1:3, 1)
addDiffLengh(1, 1:3)
```
does not behave in the way we would expect in `R`. In the first case we are adding `1` to the vector `1:3`, and the output is the same we would obtain by doing $1:3 + 1$ in `R`. But, in the second case, we are adding `1:3` to a vector of length `1` and `Rcpp` does not do any recycling.

The bottom line is that we should not expect `Rcpp` code to behave like `R` code when it comes to recycling, and that `Rcpp` does not provide a direct equivalent of the `%*%` operator for matrix-matrix and matrix-vector multiplication. Hence, when doing linear algebra in `Rcpp`, we are better off using the methods offered by the `RcppArmadillo` package, as done in the following function:  
```{r}
sourceCpp(code = '
 // [[Rcpp::depends(RcppArmadillo)]]
 #include <RcppArmadillo.h>
 using namespace Rcpp;

 // [[Rcpp::export(name = "MMv_arma")]]
 arma::vec MMv_arma_I(arma::mat& A, arma::mat& B, arma::vec& y) {
   return A * B * y;
}', verbose = TRUE)
```
Before illustrating the performance of this function, let's explain what is going on in the code above:

   - `// [[Rcpp::depends(RcppArmadillo)]]` is an `Rcpp` attribute that states that our code depends on the `RcppArmadillo`
     package. The attribute makes so that `sourceCpp` will configure the build environment to compile and link against 
     the `RcppArmadillo` package.
   - `#include <RcppArmadillo.h>` includes the code from `RcppArmadillo`, notice that we don't need to include `Rcpp.h`
     as well.
   - `arma::mat` and `arma::vec` indicate matrices and vectors of doubles defined in the `arma` namespace of the
     the `Armadillo` library. Here these are passed by reference using `arma::mat&` and `arma::vec&`, 
     hence no copy is produced. From the text output of `sourceCpp` we see that the original `SEXP` objects
     are converted to `arma::mat` objects via the `Rcpp::traits::input_parameter` mechanism. 
     The latter is beyond the scope of this course, but the point is that the conversion 
     from `R` to `arma` objects is handled automatically by `RcppArmadillo`. Conversions in the opposite direction
     are handled via `Rcpp::wrap`.

Let us now compare our `Armadillo` implementation with the plain `R` version:
```{r}
d <- 1e3
A <- matrix(rnorm(d^2), d, d)
B <- matrix(rnorm(d^2), d, d)
y <- rnorm(d)

max(abs(A %*% B %*% y - MMv_arma(A, B, y)))

library(microbenchmark)
microbenchmark(R = A %*% B %*% y, 
               Arma = MMv_arma(A, B, y))
```
The performance gains are substantial! It might be tempting to conclude that the `Armadillo` library is just so much better that `R` default linear algebra routines, but we need to understand how this performance gain was achieved. To do so, consider the following comparison: 
```{r}
microbenchmark(R = A %*% (B %*% y), 
               Arma = MMv_arma(A, B, y))
```
It seems that simply adding a pair of brackets made our `R` code much more efficient, and closer to `Armadillo`'s performance! The reason is simply that the code `A %*% B %*% y` computes the matrix product first, which is a $O(d^3)$ operation, and then multiplies that resulting matrix by the vector `y`, which is a $O(d^2)$ operation. Instead, `A %*% (B %*% y)` computes two matrix-vector multiplications, thus avoiding the $O(d^3)$ computation. The question is whether the speed-up we are getting from `Armadillo` is due to the fact that this library is automatically computing the matrix-matrix-vector product using this more efficient order of operations. To verify this, we write the following:
```{r}
sourceCpp(code = '
 // [[Rcpp::depends(RcppArmadillo)]]
 #include <RcppArmadillo.h>
 using namespace Rcpp;

 // [[Rcpp::export(name = "MMv_arma_slow")]]
 arma::vec MMv_arma_I(arma::mat& A, arma::mat& B, arma::vec& y) {
   arma::mat C = A * B;
   return C * y;
}')
```
And we compare it with the inefficient `R` implementation:
```{r}
microbenchmark(R = A %*% B %*% y,
               Arma_slow = MMv_arma_slow(A, B, y))
```
Now the performance is comparable, hence the efficiency of the `Armadillo` implementation is mostly due to the order in which the computations are performed. The fact that we can achieve similar performance using base `R` and a pair of (well-placed) brackets might be disappointing, but notice that is remarkable that `Armadillo` is able to automatically determine the order of operations so as to optimize efficiency. In reality, `Armadillo` does more than just reordering operation, but it can combine several operations into one and reduce the need for temporary objects, via delayed evaluation. This can result in faster computation and lower memory usage, relative to naive evaluation. Delayed evaluation is achieved via template meta-programming, which is beyond the scope of this course. From a user point of view, what matters is that template meta-programming allows `Armadillo` to reason about linear algebra expression at compile-time, with the aim of producing code that is tailored to each mathematical expression. See `Armadillo`'s [website](http://arma.sourceforge.net/) for more details.

Before describing some of the main functions and structures provide by `Armadillo` and `RcppArmadillo`, we cover another basic example in the next session. 

### Motivating example 2: sums of matrices

In the previous section we considered the `R` code `A %*% B %*% y` and we showed how it can be made much more efficient simply by writing it as `A %*% (B %*% y)`. Here we consider a simple linear algebra example, where improving the efficiency of a base `R` implementation is not straightforward. In particular, consider calculating ${\bf A} + {\bf B} + {\bf C} + {\bf D} + {\bf E}$, where ${\bf A}, \dots, {\bf E}$ are $d \times d$ matrices (the shape of the matrices is not important). A simple `Armadillo` function for computing such a sum is:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export(name = "mSum_arma")]]
mat mSum_i(mat& A, mat& B, mat& C, mat& D, mat& E) {
  return A + B + C + D + E;
}')
```
Let us compare it with with plain `R` code:
```{r}
A <- B <- C <- D <- E <- matrix(rnorm(1e6), 1e3, 1e3)

microbenchmark(R = Z <- A + B + C + D + E,
               arma = Z <- mSum_arma(A, B, C, D, E))
```
As we might have expected, the `Armadillo` version of the sum is faster. How was this better performance achieved? Notice that when evaluating the `A + B + C + D + E`, `R` will compute `A + B` and store the result in a temporary object, which we call `ApB`. Then, it will compute `ApB + C` and store the temporary result in another matrix, and so on. Hence, the four matrix sums will result in the allocation of four matrices, three of these are temporary and will be discarded, the last one will be stored as `Z`. An efficient implementation would avoid allocating memory for three temporary matrices, but would simply allocate memory for `Z` and then do:
```{r, eval = FALSE}
for(i in 1:d)
  for(j in 1:d)
    Z[i, j] <- A[i, j] + B[i, j] + C[i, j] + D[i, j] + E[i, j]  
```
but doing few floating point operations within a nested loop is of course hopelessly inefficient in `R`. The question is whether `Armadillo` has been able to work out that this is the best way of framing of computation. To investigate whether the performance of `Armadillo` could be attributed to such compile-time optimization, we write a version of `mSum_arma` which explicitly creates temporary matrices:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export(name = "mSum_arma_slow")]]
mat mSum_i(mat& A, mat& B, mat& C, mat& D, mat& E) {
  mat T1 = A + B;
  mat T2 = T1 + C;
  mat T3 = T2 + D;
  return T3 + E;
}')
```
We then add this version to the comparison:
```{r}
microbenchmark(R = Z <- A + B + C + D + E,
               arma = Z <- mSum_arma(A, B, C, D, E), 
               arma_slow = Z <- mSum_arma_slow(A, B, C, D, E))
```
The performance of the second `Armadillo` version is close to that of the `R` code, hence we can conclude that the efficiency of the first `Armadillo` version is attributable to the fact that it avoids creating temporary matrices.

To conclude this example, consider its extension to the problem of computing ${\bf A} + {\bf B} + \cdots$, where the number of matrices involved in the sum is arbitrary (that is, not fixed to five as above). If the matrices are contained in a `list`, for example:
```{r}
nmat <- 10
mats <- lapply(1:nmat, function(.nouse) matrix(rnorm(1e6), 1e3, 1e3))
```
then one compact way of calculating their sum in `R` is `Reduce("+", M)`. Can we modify our `RcppArmadillo` code to handle any number of matrices? One way of doing this so that the compile-time efficiency tricks of `Armadillo` are exploited is based on some text editing. In particular, consider the following function:
```{r}
getMatSumFun <- function(n){
  args <- paste0("mat& ", paste0("A", 1:n), ",", collapse = ' ')
  args <- substr(args, 1, nchar(args)-1)

  ret <- paste0(paste0("A", 1:n), " +", collapse = ' ')
  ret <- substr(ret, 1, nchar(ret)-2)

  funBody <- paste0("
   // [[Rcpp::depends(RcppArmadillo)]]
   #include <RcppArmadillo.h>
   using namespace arma;

   // [[Rcpp::export(name = \"mSum_arma_list\")]]
   mat mSum_i(", args, "){
   return ", ret, ";
  }")

  return(funBody)
}
```
Given `n`, this function creates the text for an `RcppArmadillo` function which sums up `n` matrices, in particular:
```{r}
armaCode <- getMatSumFun(nmat)
cat(armaCode)
```
Writing down the sum explicily guarantees that `Armadillo` will avoid creating temporary matrices. We can now compare the `R` and `Armadillo` solutions:
```{r}
sourceCpp(code = armaCode)

microbenchmark(R = Reduce("+", mats), arma = do.call("mSum_arma_list", mats))
```

### Basic `RcppArmadillo` usage

Here we describe some commonly used features of `RcppArmadillo` and `Armadillo`. The `Armadillo` library is quite extensive, hence we refer to its [documentation](http://arma.sourceforge.net/docs.html) for more details. The most commonly used object classes are:

   - `Mat<type>` which represents dense matrices, with element types defined via the template argument `type`. In the 
      examples above we used `mat` which is an alias or typedef for `Mat<double>`. 
   - `Col<type>` and `Row<type>` which represent column and row vectors. Shortcuts for `Col<double>` and `Row<double>` are 
     `colvec` (or just `vec`) and `rowvec`.  
     
`Armadillo` provides also 3D matrices, sparse matrices and fields (matrices whose elements can be arbitrary objects, such as matrices), which we will not use here. In order to effectively use an `Armadillo` object class, we need to know (or look in the [documentation](http://arma.sourceforge.net/docs.html) for):

   1. the **constructor** types available for such class. For example, a matrix of doubles `A` could be construted using 
      `mat A(10, 5)` or `mat A(10, 5, fill::zeros)`, for example. The difference being that in the first case the 
      entries of the $10 \times 5$ matrix `A` are arbitrary, while in the second they are set to 0. 
   2. The member **functions and variables** of that class. These are accessed using `object.member_function()` or 
      `object.variable`. For example, if `y` is a `Col<float>` vector, the `y.n_elem` gives its lengths while `y.ones()` 
      sets all its elements to 1. The [docs](http://arma.sourceforge.net/docs.html) clearly explain which member functions
      are available for each object class.
   3. How the elements of the objects can be accessed. This is documented as part of the "Member Functions & Variables" 
      section the `Armadillo` docs. An element $ij$ of a matrix `A` of class `Mat` can be accessed via `A(i-1,j-1)`, as 
      indexing starts at zero. It can also be accessed via `A.at(i-1,j-1)` which is faster but more dangerous as no 
      bounds check is performed (hence we could read/write outside the memory allocated to `A`).
   4. The available **operators** and **functions**. Operators are overloaded so if, for example, `A` and `B` are 
      matrices while `s` is a scalar, then `A * B` will calculate the standard matrix product while `A * s` will 
      multiply each element of `A` by `s`. Several element-wise functions are provided (e.g., `exp(A)`, `sqrt(A)`, ...), 
      as well as standard decomposition (e.g, Cholesky, LU, ...) and solvers for linear systems. 
      
Some guidance on how standard linear algebra operations in `R` can be translated to `RcppArmadillo` is provided [here](http://arma.sourceforge.net/docs.html). Before moving to a more involved example, we clarify the relation between `Armadillo` objects and `Rcpp` matrices and vectors. Conversion from `Armadillo` object to `Rcpp` object is performed using `Rcpp::wrap`, for example:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export(name = "getIntMat")]]
Mat<int> tmpFun_i(int nr, int nc) {
  Mat<int> MA(nr, nc);
  Rcpp::IntegerMatrix MR = Rcpp::wrap(MA);
  MR[0, 0] = 1;
  return MA;
}')

getIntMat(3, 3)
```
Here we are creating a $3 \times 3$ un-initialized integer matrix in `Armadillo` and we are converting it to an `Rcpp::IntegerMatrix` using `wrap`. It is interesting to notice that here `Rcpp::wrap` is copying `MA`, which is demostrated by the fact that changing the top-left element of `MR` does not entail changing `MA`. The following version avoids any copy: 
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export(name = "getIntMat_noCopy")]]
Rcpp::IntegerMatrix tmpFun_i(int nr, int nc) {
  Rcpp::IntegerMatrix MR(nr, nc);
  Mat<int> MA(MR.begin(), MR.nrow(), MR.ncol(), false);
  MA(0, 0) = 1;
  return MR;
}')

getIntMat_noCopy(3, 3)
```
Let's see how this worked step by step:

   - `Rcpp::IntegerMatrix MR(nr, nc)` here we **allocated memory in R** for the matrix `MR`;
   - `Mat<int> MA(MR.begin(), MR.nrow(), MR.ncol(), false)` here we used the advanced `Armadillo` 
      [constructor](http://arma.sourceforge.net/docs.html#adv_constructors_col), which constructs a `Mat<int>` 
      matrix `MA` by reusing the memory allocated to `MR`. Hence no copy is made and changes in `MA` leads to 
      changes in `MR`.
   - We return the `IntegerMatrix` `MR`. Given that its memory is allocated in `R`, no copy is made when
     the object is returned. In particular, recall that `sourceCpp` will create a wrapper around our `C++` function, 
     and the wrapper will contain the line `Rcpp::wrap(tmpFun_i(nr, nc))`, as you can verify by calling 
     `sourceCpp` with `verbose = TRUE`. In `getIntMat`, `Rcpp::wrap` was taking a copy, because the output 
     of `tmpFun_i` does not use `R` memory, while this is avoided in `getIntMat_noCopy`. 
     
In summary `getIntMat` allocates memory outside `R` via `Mat<int> MA(nr, nc);`, than takes one copy when we do `Rcpp::wrap(MA)` and another copy when `Rcpp::wrap` is called within the wrapper written by `sourceCpp`. In contrast, 
`getIntMat_noCopy` allocates `R` memory via `Rcpp::IntegerMatrix MR(nr, nc);` and does no take any copy. This makes the second version much more efficient, in fact: 
```{r}
d <- 1e3
microbenchmark(copy = getIntMat(d, d), noCopy = getIntMat_noCopy(d, d))
```
The bottom line is that, when we want to manipulate large matrices or vectors in `RcppArmadillo`, it is often more efficient to: 

   1. (a) allocate memory in `R` and then pass the objects by reference (e.g., via `mat&`, `vec&`, ...) to the `Rcpp` function 
      or (b) to allocate `R` memory directly in `Rcpp` via one of its data structure (e.g., `NumericMatrix`, `NumericVector`, ...);
   2. wrap the `R` object into `Armadillo` objects using an advanced constructor, such as the one we have see above;
   3. manipulate the `Armadillo` objects using the operators and functions provided by `Armadillo`;
   4. return an `Rcpp` data structure to avoid copies being made by `Rcpp::wrap`.
   
The advance constructor provides a way of converting `Rcpp` objects into `Armadillo` objects, without taking a copy. Often talking a copy is acceptable, in which case it is more convenient to perform the conversion using the `Rcpp::as<type>()` function, where the template argument `type` can be an `Armadillo` object class such as `Mat<int>`, `Row<double>` and so on. An example is:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export(name = "matVettArma")]]
Col<double> tmpFun_i(Rcpp::NumericMatrix X, Rcpp::NumericVector y) {
  Mat<double> Xa = Rcpp::as<Mat<double>>(X);
  Col<double> ya = Rcpp::as<Col<double>>(y);
  return Xa*ya;
}')

matVettArma(matrix(1, 3, 3), 1:3)
```
which performs matrix-vector multiplication in `Armadillo`. Notice that `Rcpp::as` generally takes a copy of the object being converted.

### References

- Armadillo project: http://arma.sourceforge.net/

- Conrad Sanderson and Ryan Curtin. Armadillo: a template-based C++ library for linear algebra.
  Journal of Open Source Software, Vol. 1, pp. 26, 2016.
  
- Dirk Eddelbuettel and Conrad Sanderson, "RcppArmadillo: Accelerating R with high-performance
  C++ linear algebra", Computational Statistics and Data Analysis, 2014, 71, March, pages 1054-
  1063, http://dx.doi.org/10.1016/j.csda.2013.02.005. )