---
title: 3. Example on multivariate KDE with RcppArmadillo
weight: 3
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

### Introduction

To provide a more challeging example illustrating the capabilities of `RcppArmadillo`, here we show how to perform multivariate kernel density estimation (k.d.e.) using this library. In particular, let $\bf x_1^o, \dots, \bf x_n^o$ be $d$-dimensional vectors, sample from the density $\pi(\bf x)$. A k.d.e. estimate of $\pi({\bf x})$ is
$$
\hat{\pi}_{\bf H}({\bf x}) = \frac{1}{n} \sum_{i=1}^n \kappa_{\bf H}({\bf x} - {\bf x}_i^o),
$$
where $\kappa_{\bf H}$ is a kernel with positive definite bandwidth matrix ${\bf H}$. Here we assume that $\kappa_{\bf H}$ is a multivariate Gaussian kernel, with covariance matrix $\kappa_{\bf H}$. Further, the data is generated the mixture of warped Gaussians of [Fasiolo et al. (2018)](https://arxiv.org/pdf/1611.06874.pdf). The details of this density are unimportant here, what matters is that the example is defined in $d>2$ dimensions and, under such density $\pi({\bf x})$, the joint distribution of the first two dimensions is far from $Gaussian$ the remaining dimensions are standard i.i.d. Gaussian variables.

Samples from this density can be generated using the following functions:
```{r}
rBanana <- function(n, d, a, b, shi1, shi2){
  out <- matrix(rnorm(d*n), n, d);
  out[ , 1] <- out[ , 1] * a
  out[ , 2] <- out[ , 2] - b * (out[ , 1]^2 - a^2)
  out[ , 1] <- out[ , 1] + shi1
  out[ , 2] <- out[ , 2] + shi2
  return( out )
}

rBanMix <- function(n, d, w, a, b, shi1, shi2){
  nmix <- length( a );
  m <- floor( n * w );
  m[1] <- m[1] + n - sum(m);
  m <- round(m);
  out <- lapply(1:nmix, function(.ii)
                         rBanana(m[.ii], d, a[.ii], b[.ii], shi1[.ii], shi2[.ii])
               )
  out <- do.call("rbind", out)
  return( out );
}
```
which we use here to generate $10^3$ variables in $5$ dimensions:
```{r}
d <- 5;
bananicity = c(0.2, -0.03, 0.1, 0.1, 0.1, 0.1)
sigmaBan <- c(1, 6, 4, 4, 1, 1)
banShiftX <- c(0, 0, 7, -7, 7, -7)
banShiftY <- c(0, -5, 7, 7, 7.5, 7.5)
nmix <- length(bananicity)
bananaW = c(1, 4, 2.5, 2.5, 0.5, 0.5)
bananaW <- bananaW / sum(bananaW)

x <- rBanMix(1e3, d, bananaW, sigmaBan, bananicity, banShiftX, banShiftY)

pairs(x)
```
The plot demonstrate that only the first tow dimensions of $\bf x$ are far from Gaussian. The next section presents a first attempt at estimating the density using `R` and `Rcpp`.

### An R-based and a dumb `Rcpp` solution

We start by creating a generic `R` function for kernel density estimation in `R`:
```{r}
kdeR <- function(dker, y, x, H){
  n <- nrow(x)
  m <- nrow(y)
  out <- numeric(m)
  for(ii in 1:n){
    out <- out + dker(y, x[ii, ], H)
  }
  out <- out/n
  return(out)
}
```
This function works as follows:

   - `y` is the $m \times d$ matrix at points at which we want to evaluate the k.d.e.;
   - `x` is the $n \times d$ matrix of original samples;
   - `H` is the bandwith matrix. 

We are now ready to evaluate our k.d.e. a grid of $10^4$ points along the first two dimensions:
```{r}
l <- 100
x1 <- seq(-20, 20, length.out = l);
x2 <- seq(-15, 15, length.out = l);
grd <- as.matrix(cbind(expand.grid(x1, x2), matrix(0, l^2, d-2)))

library(mvtnorm)
dns <- kdeR(dmvnorm, grd, x, diag(d))
```
Where we use the `dmvnorm` function from the `mvtnorm` package to evaluate the Gaussian kernel and we set $\bf H$ simply to the identity matrix. We plot the `k.d.e.`:
```{r, fig.align='center', fig.height=4, fig.width=5}
library(ggplot2)
library(viridis)
ggplot(data = data.frame(x = grd[ , 1], y = grd[ , 2], z = dns),
       mapping = aes(x = x, y = y, z = z, fill = z)) +
       geom_raster() + geom_contour() + scale_fill_viridis_c()
```

Can we speed the evaluation of the $k.d.e.$ up using `RcppArmadillo`? We start with a "lazy" version, where we replace `kdeR` with:
```{r}
library(Rcpp)
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

// [[Rcpp::export(name = "kdeLazy")]]
Rcpp::NumericVector kde_i(Rcpp::Function dker, mat& y, mat& x, mat& H) {
  unsigned int n = x.n_rows;
  unsigned int m = y.n_rows;
  vec out(m, fill::zeros);
  for(int ii = 0; ii < n; ii++){
   out += Rcpp::as<vec>(dker(y, x.row(ii), H));
  }
  out /= n;
  return Rcpp::wrap(out);
}')
```
Which can be used identically to `kdeR` and produces the same estimates:
```{r, fig.align='center', fig.height=4, fig.width=4}
plot(kdeR(dmvnorm, x, x, diag(d)),
     kdeLazy(dmvnorm, x, x, diag(d)))
abline(0, 1, col = 2)
```
Let us comparing the performance of `kdeR` and `kdeLazy`:
```{r}
library(microbenchmark)
microbenchmark(kdeR = kdeR(dmvnorm, x, x, diag(d)),
               kdeLazy = kdeLazy(dmvnorm, x, x, diag(d)))
```
It turns out that the `Rcpp` version is not faster. This demonstrates that writing part of the code in `Rcpp` does automatically lead to performance gains. In particularly our `Rcpp` code:

   - calls the `dmvnR` function to compute the Gaussian kernel. We cannot expect an `R` function to be      any faster when call from `C++` rather than from `R`!
   - the call to `Rcpp::as<vec>` in the main `for` loop makes a copy at each call, which is
     inefficient.
     
The next section proposes a "proper" `RcppArmadillo` solution for multivariate kernel density estimation.


### A better `RcppArmadillo` solution

We start by developing an efficient `RcppArmadillo` function for evaluating the multivariate normal density. Before thinking about its implementation in software, we have to think about what is the most efficient way of computing it numerically. The formula for the log-density is
$$
\log \phi({\bf x}, \mu, {\bf H}) = -\frac{1}{2}({\bf x} - \mu)^T{\bf H}^{-1}({\bf x} - \mu) - \frac{d}{2} \log (2\pi) - \frac{1}{2}\log\text{det}({\bf H}). 
$$
Now, for the purpose of maximizing efficiency, we have to think about how to numerically compute the quadratic form and the determinant. Let's start with the first. 

Remember that, in our k.d.e. application, ${\bf x}_1, \dots, {\bf x}_m$ are the points at which we want to evaluate the k.d.e. and $\mu_1 = {\bf x}^o_1, \dots, \mu_n = {\bf x}^o_n$ represent the centers of the kernel densities. Hence, for any fixed value of $\mu$, we will have to evaluate the quadratic form $m$ times. For fixed $\mu$, we could considering two approaches:

   A) calculate the inverse of ${\bf H}$ upfront and then calculate 
      $({\bf x}_i - \mu)^T{\bf H}^{-1}({\bf x}_i - \mu)$, for $i = 1, \dots, m$;

   B) solve the linear system ${\bf H}^{-1}({\bf x}_i - \mu) = {\bf z}_i$ and then do 
      $({\bf x}_i - \mu)^T{\bf z}_i$, for $i = 1, \dots, m$;
      
Approach A seems quite appealing: we pay an upfront $O(d^3)$ cost to get ${\bf H}^{-1}$ and the we pay $O(d^2)$ cost $m$ times to evaluate the quadratic forms. So, for $d << m$, the dominant cost is $O(md^2)$. Approach B seems more wasteful: solving the linear system by Gaussian elimination will cost $O(d^3)$ operations, so doing it $m$ times brings the dominant cost to $O(md^3)$. However, here it is essential to take into account the fact that $\bf H$ is positive definite, which means that approach B can be implemented much more efficiently. In particular, let ${\bf L}{\bf L}^T = {\bf H}$ be the Cholesky decomposition of $\bf H$. Then ${\bf H}^{-1}({\bf x}_i - \mu) = ({\bf L}^{-1})^T{\bf L}^{-1}({\bf x}_i - \mu)$ can be computed by:

   1. solving ${\bf L}^{-1}({\bf x}_i - \mu) = {\bf z}_i$ by forward-substitution 
      (recall that $L$ is lower triangular). This has cost $O(d^2)$.
   2. solving $({\bf L}^{-1})^T {\bf z}_i = \tilde{\bf z}_i$ by back-substitution. Again $O(d^2)$.
   3. computing $({\bf x}_i - \mu)^T \tilde{\bf z}_i$ which is $O(d)$.
   
Hence, the cost of approach B is $O(m d^2)$ as well, but has the advantage of avoiding matrix inversion which is generally less inaccurate than solving a linear system.

The approach we are going to follow is based on the observation that $({\bf x}_i - \mu)^T {\bf H}^{-1}({\bf x}_i - \mu) = {\bf z}_i^T{\bf z}_i$, hence once we have obtained ${\bf z}_i$ by forward-substitution as in step 1 above, we just need to calculate its squared norm at cost $O(d)$. Having obtained the cholesky decomposition of $\bm H$, computing its log-determinant is very cheap because $\log\text{det}({\bf H}) = \log\text{det}({\bf L}{\bf L}^T) = 2\log\text{det}({\bf L}) = 2\log\text{trace}({\bf L})$.
Before lookign at the `Armadillo` function will evaluate the Gaussian log-density, which we call `dmvnInt`, lets look at the function that will be accessed at R level:
```{r engine = 'Rcpp', eval = FALSE}
Rcpp::NumericVector kde_i(mat& y, mat& x, mat& H) {
  unsigned int n = x.n_rows;
  unsigned int m = y.n_rows;
  vec out(m, fill::zeros);
  mat cholDec = chol(H, "lower");
  for(int ii = 0; ii < n; ii++){
   out += dmvnInt(y, x.row(ii), cholDec);
  }
  out /= n;
  return Rcpp::wrap(out);
```
This is almost identical to the function that we have used before, the main differences being the following lines:

   - `mat cholDec = chol(H, "lower");` here we are computing the Cholesky decomposition of $\bf H$ 
     and we are storing the lower-triangular factor in `cholDec`;
   - `out += dmvnInt(y, x.row(ii), cholDec);` evaluates a multivariate normal density, with mean vector
      $\mu = {\bf x}_i^0$ and covariance $\bf H$, at ${\bf x}_1, \dots, {\bf x}_m$. Instead of passing
      $\bf H$ to `dmvnInt`, we pass it directly its lower-triangular factor. We assume that the output 
      `dmvnInt` is an `arma::vec` which can be accumulated in `out` without calling `Rcpp::as<vec>`.
      
Then we examine the function for evaluating the multivariate Gaussian density:
```{r engine = 'Rcpp', eval = FALSE}
vec dmvnInt(mat & X, const rowvec & mu, mat & L)
{
  unsigned int d = X.n_cols;
  unsigned int m = X.n_rows;
  
  vec D = L.diag();
  vec out(m);
  vec z(d);
  
  double acc;
  unsigned int icol, irow, ii;
  for(icol = 0; icol < m; icol++)
  {
    for(irow = 0; irow < d; irow++)
    {
     acc = 0.0;
     for(ii = 0; ii < irow; ii++) acc += z.at(ii) * L.at(irow, ii);
     z.at(irow) = ( X.at(icol, irow) - mu.at(irow) - acc ) / D.at(irow);
    }
    out.at(icol) = sum(square(z));
  }

  out = exp( - 0.5 * out - ( (d / 2.0) * log(2.0 * M_PI) + sum(log(D)) ) );

  return out;
}
```
the important lines are:

   - `vec dmvnInt(mat & X, const rowvec & mu, mat & L)` we return an `arma::vec` and all 
     inputs are passed by reference;
   - `vec D = L.diag(); vec out(m); vec z(d);` we extract the main diagonal of `L`, we defined the
      $m$-dimensional vector `out` will contain the density values at ${\bf x}_1, \dots, {\bf x}_m$ 
      and "working" $d$-dimensional vector `z` which we will use in the main loop;
   - `for(icol = 0; icol < m; icol++)` here we loop across ${\bf x}_1, \dots, {\bf x}_m$. 
     We calculate ${\bf z}_i$ using the nested loop (next bullet point) and then
     we use `out.at(icol) = sum(square(z));` to compute and store $||{\bf z}_i||^2$.
   - `for(irow = 0; irow < d; irow++)` given ${\bf x}_i$ we get 
     ${\bf z}_i = {\bf L}^{-1}({\bf x}_i - \mu)$ by forward-substitution.
     
Let use now compile and load using `sourceCpp`:
```{r}
sourceCpp(code = '
// [[Rcpp::depends(RcppArmadillo)]]
#include <RcppArmadillo.h>
using namespace arma;

vec dmvnInt(mat & X, const rowvec & mu, mat & L)
{
  unsigned int d = X.n_cols;
  unsigned int m = X.n_rows;
  
  vec D = L.diag();
  vec out(m);
  vec z(d);
  
  double acc;
  unsigned int icol, irow, ii;
  for(icol = 0; icol < m; icol++)
  {
    for(irow = 0; irow < d; irow++)
    {
     acc = 0.0;
     for(ii = 0; ii < irow; ii++) acc += z.at(ii) * L.at(irow, ii);
     z.at(irow) = ( X.at(icol, irow) - mu.at(irow) - acc ) / D.at(irow);
    }
    out.at(icol) = sum(square(z));
  }

  out = exp( - 0.5 * out - ( (d / 2.0) * log(2.0 * M_PI) + sum(log(D)) ) );

  return out;
}

// [[Rcpp::export(name = "kdeArma")]]
Rcpp::NumericVector kde_i(mat& y, mat& x, mat& H) {
  unsigned int n = x.n_rows;
  unsigned int m = y.n_rows;
  vec out(m, fill::zeros);
  mat cholDec = chol(H, "lower");
  for(int ii = 0; ii < n; ii++){
   out += dmvnInt(y, x.row(ii), cholDec);
  }
  out /= n;
  return Rcpp::wrap(out);
}')
```
Notice that we are not exporting `dmvnInt`, which is used only internally by `kde_i`. Let's test whether our `Armadillo` version works:
```{r, fig.align='center', fig.height=4, fig.width=4}
plot(kdeR(dmvnorm, x, x, diag(d)),
     kdeArma(x, x, diag(d)))
abline(0, 1, col = 2)
```
It seems so! We now compare the `Armadillo` and the `R` version in terms of speed:
```{r}
library(microbenchmark)
microbenchmark(R = kdeR(dmvnorm, x, x, diag(d)),
               Armadillo = kdeArma(x, x, diag(d)))
```
This is definitely better than the previous version! Notice that the `R` version (`kdeR`) is not too inefficient here, because at each iteration of the main `for` loop the `dmvnorm` evaluates a Gaussian density at $10^3$ data points. If we reduce that number of points at which the k.d.e. is evaluated, the performance gains increase. For instance:
```{r}
library(microbenchmark)
microbenchmark(R = kdeR(dmvnorm, x[1:10, ], x, diag(d)),
               Armadillo = kdeArma(x[1:10, ], x, diag(d)))
```
This is attributable to the fact that we are computing the Cholesky decomposition of $\bf H$ only once, rather than $n$ times, to the fact that `for` loops are much quicker in `C++` that in `R` and, hopefully, that our code for computing the multivariate normal density is faster than that used by `dmvtnorm`.

### References

- Armadillo project: http://arma.sourceforge.net/

- Conrad Sanderson and Ryan Curtin. Armadillo: a template-based C++ library for linear algebra.
  Journal of Open Source Software, Vol. 1, pp. 26, 2016.
  
- Dirk Eddelbuettel and Conrad Sanderson, "RcppArmadillo: Accelerating R with high-performance
  C++ linear algebra", Computational Statistics and Data Analysis, 2014, 71, March, pages 1054-
  1063, http://dx.doi.org/10.1016/j.csda.2013.02.005. )
  
- Fasiolo, M., de Melo, F.E. and Maskell, S., 2018. Langevin incremental mixture importance sampling.    Statistics and Computing, 28(3), pp.549-561.