---
title: 4. Exercise on local polynomial regression
weight: 4
output:
  blogdown::html_page:
    toc: true
---

<style>
body {
text-align: justify}
</style>

### Smoothing by local polynomial regression

Consider the following data set on solar [electricity production from Sidney, Australia](https://www.ausgrid.com.au/Industry/Our-Research/Data-to-share/Solar-home-electricity-data):
```{r, message=FALSE}
load("solarAU.RData")
head(solarAU)
```
The variables are:

  - `prod` total production from 300 homes;
  - `toy` time-of-year, going from 0 to 1 (00:00 on Jan 1st to 23:30 on 31st Dec);
  - `tod` time-of-day, taking value in 0, 1, 2, ..., 47 (00:00, 00:30, ..., 23:30);
  
The interest is modelling the production as a function of the `tod` and `toy`. We will be working with log-production, which is less skewed:
```{r}
solarAU$logprod <- log(solarAU$prod+0.01)
```
We added $0.01$ to avoid getting `-Inf` when the production is zero. Let's look at the log-production as a function of the two covariates:
```{r, message = FALSE}
library(ggplot2)
library(viridis)
ggplot(solarAU,
       aes(x = toy, y = tod, z = logprod)) +
       stat_summary_hex() +
       scale_fill_gradientn(colours = viridis(50))
```
As expected, there are more hours of daylight in Austral winter, hence more production occurs during that period of the year. 

Now, we aim at modelling the expected log-production, $y$, as a function of ${\bf x} = \{\text{tod}, \text{toy}\}$. That is, we want a model for $\mathbb{E}(y|{\bf x})$. We start with a simple polynomial regression model:
$$
\mathbb{E}(y|{\bf x}) = \beta_0 + \beta_1\text{tod} + \beta_2\text{tod}^2 + \beta_3\text{toy} + \beta_4\text{toy}^2 = \tilde{\bf x}^T\beta.
$$
where $\tilde{\bf x} = \{\text{tod}, \text{tod}^2, \text{toy}, \text{toy}^2\}$. It is quite simple to fit this model in `R`:
```{r}
fit <- lm(logprod ~ tod + I(tod^2) + toy + toy^2, data = solarAU) 
```

**Q1 start** Use `RcppArmadillo` to fit a linear regression model, that is to solve the minimization problem:
$$
\hat{\beta} = \underset{\beta}{\text{argmin}} ||{\bf y} - {\bf X}\beta||^2,
$$
where, in this example, the model matrix is given by:
```{r}
X <- with(solarAU, cbind(1, tod, tod^2, toy, toy^2))
```
Think about what numerical approach you will use (e.g., cholesky or QR decomposition) and compare the speed of your implementations with `lm` (of course, first you need to check whether your function gives correct results!).
**Q1 end**

As the following plots show, the polynomial fit is not great:
```{r, fig.align='center', fig.height=4, fig.width=10}
library(gridExtra)

solarAU$fitPoly <- fit$fitted.values

pl1 <- ggplot(solarAU,
              aes(x = toy, y = tod, z = fitPoly)) +
       stat_summary_hex() +
       scale_fill_gradientn(colours = viridis(50))

pl2 <- ggplot(solarAU,
              aes(x = toy, y = tod, z = logprod - fitPoly)) +
       stat_summary_hex() +
       scale_fill_gradientn(colours = viridis(50))
grid.arrange(pl1, pl2, ncol = 2)
```
In particular, the second plot shows a clear non-linear pattern in the residuals. To improve our fit, we make our linear regression model locally adaptive by adopting a local least regression approach (see, e.g., Hastie et al, 2009, section 6.1.2 and 6.3). This consists in letting the estimated regression coefficients depend on $\bf x$, so that $\hat{\beta} = \hat{\beta}({\bf x})$. To achieve this, for a fixed value ${\bf x}_0$, we find $\hat{\beta}({\bf x}_0)$ by minimizing the following objective:
$$
\hat{\beta}({\bf x}_0) = \underset{\beta}{\text{argmin}} 
\sum_{i=1}^n \kappa_{\bf H}({\bf x}_0-{\bf x}_i) (y_i - \tilde{\bf x}_i^T\beta)^2,
$$
where $\kappa_{\bf H}$ is a density kernel with positive definite bandwidth matrix $\bf H$. Given that $\kappa_{\bf H}({\bf x}_0-{\bf x}_i) \rightarrow 0$ as $||{\bf x}_0-{\bf x}_i|| \rightarrow \infty$, we have that $\hat{\beta}({\bf x}_0)$ will depend more strongly on the data points close to ${\bf x}_0$ than on those far from it. 

It is quite easy to create a function that does this in `R`:
```{r}
library(mvtnorm)
lmLocal <- function(y, x0, X0, x, X, H){
  w <- dmvnorm(x, x0, H)
  fit <- lm(y ~ -1 + X, weights = w)
  return( t(X0) %*% coef(fit) )
}
```
where we are using the Gaussian kernel. Notice that this requires re-estimating the model for each location ${\bf x}_0$ we are interested in. In our case:
```{r}
nrow(solarAU)
```
so we need to fit the local regression model more than $17$ thousands times to get an estimate of $\mathbb{E}(y|{\bf x}_0)$ at each observed location. Hence, we test our local model on a sub-sample of 2000 data points: 
```{r}
n <- nrow(X)
nsub <- 2e3
sub <- sample(1:n, nsub, replace = FALSE)

y <- solarAU$logprod
solarAU_sub <- solarAU[sub, ]
x <- as.matrix(solarAU[c("tod", "toy")])
x0 <- x[sub, ]
X0 <- X[sub, ]
```
We can now obtain estimates at each subsampled location (this might take a minute or two):
```{r}
predLocal <- sapply(1:nsub, function(ii){
  lmLocal(y = y, x0 = x0[ii, ], X0 = X0[ii, ], x = x, X = X, H = diag(c(1, 0.1)^2))
})
```
Notice that we are setting $\bf H$ to be a diagonal matrix with hand-chosen diagonal entries. Let's look at the fit.
```{r, fig.align='center', fig.height=4, fig.width=10}
solarAU_sub$fitLocal <- predLocal

pl1 <- ggplot(solarAU_sub,
       aes(x = toy, y = tod, z = fitLocal)) +
       stat_summary_hex() +
       scale_fill_gradientn(colours = viridis(50))

pl2 <- ggplot(solarAU_sub,
       aes(x = toy, y = tod, z = logprod - fitLocal)) +
       stat_summary_hex() +
       scale_fill_gradientn(colours = viridis(50))

grid.arrange(pl1, pl2, ncol = 2)
```
The left plot looks quite similar to the data (first plot of this document) and the residual plot shows no clear residual pattern. However, our code is very slow, so:

**Q2 start** To speed up the local least squares fit, implement it in `RcppArmadillo` and compare your solutions with the `R` code above in terms of speed (and correctness). If you wish, you can re-use the code we have described in a previous chapter for evaluating the multivariate normal density. **Q2 end**

**Q3 start** Above, we have chosen the bandwidth matrix $\bf H$ manually. Once you have created an `RcppArmadillo` function for local linear regression, set up a cross-validation routine for tuning the bandwidth matrix automatically. **Q3 end**

### References

- Hastie, T., Tibshirani, R. and Friedman, J., 2009. The elements of statistical learning: data mining, inference, and prediction (Second Edition). Springer.








